---
title: "M4: ctmm-Based Movement Metrics: Speed and Distance"
subtitle: "Day 2 — Movement Metrics"
author: "Animal Movement Analyses for Ecologists | SYSU Shenzhen, July 2026"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    toc_collapse: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Created: 2026-06-30* | *Last updated: `r Sys.Date()`* 

<a href="https://github.com/Animal-Movements/China_2026.git" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#151513; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

---

## Learning Objectives

By the end of this module you will be able to:

1. Estimate the mean speed of an animal from its tracking data.
2. Perform a meta-analysis to estimate the mean speed of a population.
3. Calculate instantaneous speeds and derive distances from speed estimates.

**Packages used:** `ctmm`

---

## Background

Estimating the movement speed of an individual animal and its distance traveled can provide insights into its behavior and activity (energetics). Ideally, we would need very fine sampling to get good, realistic estimates of distance traveled based on an individual's tracked movement path, from which we can calculate its speed. Of course, this is not always feasible and there are often many logistical constraints for using a short sampling interval, as well as [trade-offs with sampling duration](https://doi.org/10.1002/ece3.73755).

Conventional straight-line distance (SLD) can be easily calculated as $\sqrt{(\Delta x^2+\Delta y^2)}$, but this is heavily biased. SLD will always underestimate the true distance traveled, unless an animal moves in a perfectly straight line. And, in cases where the animal does not move much relative to the GPS location error (*more to come*) of its tracking device, SLD can greatly overestimate the distance traveled. This also impacts our speed estimates. Here, we will use the [continuous-time speed and distance (CTSD)](https://doi.org/10.1186/s40462-019-0177-1) estimator to get more accurate mean speed, instantaneous speeds, and distance estimates based on the CTMM framework.

## Data preparation

Like in the previous module, let's first import our data, center the projection so that North points up, and then visualize our data. We will use the GPS-collared African buffalo (*Syncerus caffer*) data set again, but we will focus just on the first individual, Cilla.

### Import and Visualize Data

```{r import_data}

# Load ctmm package
library(ctmm)

# Load buffalo dataset from ctmm
data(buffalo)

# Projection with North = up
projection(buffalo) <- median(buffalo)  # center on geometric median of locations

# Let's consider just first buffalo (Cilla)
DATA <- buffalo[[1]]

# Plot Cilla's tracks
plot(DATA, col = color(DATA, by = "time"),  # color by time
     main = "Cilla's tracks colored by time")

```

### Model Selection

```{r ctmm.select, eval = FALSE}

# Model selection
## Guesstimate autocorrelation parameters
GUESS <- ctmm.guess(data = DATA, interactive = FALSE)  
FIT <- ctmm.select(data = DATA, CTMM = GUESS)  # fit and select movement models
# save(FIT, file = "Data/Cilla.rda")  # save results

```

```{r ctmm.select2}

# Load model fits from ctmm.select
load("Data/cilla.rda")

# Remember our handy units operator?
# help(`%#%`)  # conversion to SI units

1 %#% 'day'  # 1 day in seconds
1 %#% 'year'  # 1 year in seconds

```

For the sake of time, we will demonstrate speed estimation on only the first week of data. This is where handy `%#%` unit conversion operator can be especially helpful. We will use it to subset our data to the first week, by having it calculate the number of seconds in 1 week and keeping everything in our data that is equal or less than that.

```{r ctmm.select3}

# For the sake of time, we will demonstrate only on the first week of data
DATA <- DATA[DATA$t <= DATA$t[1] + 1%#%'week',]  # subset to first week

# Plot without errors (easier to see the location points)
plot(DATA, col = color(DATA, by = 'time'),  # color by time
     error = FALSE, main = "First week of tracking data")
## Less data so ctmm.select should fit relatively quicker

# Re-fit to first week only
FIT <- ctmm.select(DATA, FIT, trace = 1)
summary(FIT)$name  # OUF anisotropic was the best fit model

```

## Speed and Distance Estimation

```{r speed}

# the speed estimate here is RMS Gaussian
summary(FIT)
## contains Gaussian root-mean squared (RMS) speed -- proxy of speed
## mean speed: v^- = E[v] = 1/(tn-t1)*integral of speed over time
## RMS speed: v_rms = root(E[v^2])

# calculate Gaussian (regular mean speed - not RMS)  -- should be w/in an order of magnitude from RMS speed
speed(FIT, trace = FALSE)  # keep trace = TRUE to see the progress bar
## both speeds assume Gaussian distr of locations and speeds

# non-parametric speed estimation
# "2019 Noonan Fleming Akre ... Calabrese.pdf" in Readings/Continuous_Time folder
speed(DATA, FIT, trace = FALSE)
## gives mean of mean speed
## data for species w/ vary distinct movement behaviours, speed calcs should be segmented
#### ex. birds stationary/feeding on ground VS flying around
## can also segment day and night; if not, slight underestimation during day, slight overestimation at night
## fairly insensitive to the effect of coarsening data (other methods would estimate animal to be slower for coarser sampled data)

```

### Sensitivity Analysis Case Study

```{r speed_coarsen}

# Impact of coarsening the data
SUB <- DATA
FIT.SUB <- FIT

# Remove every other time
## This coarsens the data by removing half of it, while maintaining the duration
## We are effectively sampling half as frequently (double sampling interval)
SUB <- SUB[as.logical(1:nrow(SUB)%%2),]
FIT.SUB <- ctmm.select(SUB, FIT.SUB, trace = FALSE)  # fits more quickly w/ less data
# the speed estimate here is RMS Gaussian
## won't get speed estimate if too coarse (resolution of data not high enough)

summary(FIT.SUB)  # wider CIs
# Gaussian (regular speed - not RMS)

# Mean speed
speed(FIT.SUB, trace = FALSE)  
## similar estimates to finer sampled data, but more uncertainty (larger CIs)

# non-parametric speed estimation
speed(SUB, FIT.SUB, trace = FALSE)  
## fits more slowly w/ less data (note lower DoF for speed, effective sample size for speed estimation)
## At some point where data is too coarse, the movement model might change to one that doesn't incorporate a velocity autocorrelation component
## Will get no speed estimate in that case
#########################
# repeat until data become too coarse
## Can look instead at diffusion rate if data too coarse (diffusion rate as a more reliable metric of animal movement activity)
## Can also look at location error

# keep in mind the stationary assumption of the model
# see the appendix of Noonan et al.

```

## Population Mean Speed

```{r ctmm_meta, eval = FALSE}

help('meta')  # population meta-analysis for movement parameters

```

```{r ctmm_meta2}

# Load in the fitted movement models for all 6 buffalo
load("Data/buffalo.rda")

```

```{r ctmm_meta3, eval = FALSE}

# Estimate mean speed for each buffalo
SPEEDS <- list()  # empty list to store results
for(i in 1:length(buffalo)) {
  
  # Estimate mean speed and store in the list
  SPEEDS[[i]] <- speed(buffalo[[i]], FITS[[i]])  # this can be slow
}

names(SPEEDS) <- names(buffalo)  # match names to the buffalo
# save(SPEEDS, file = "Data/buffalo_speeds.rda")  # save the results

```

```{r ctmm_meta4}

# Since it can be a slower process, let's load the saved file
load("Data/buffalo_speeds.rda")  # mean speed estimates for each buffalo

meta(SPEEDS, sort = TRUE)  # mean = population mean of mean speeds

```

## Instantaneous Speeds

```{r ctmm_instant_speed}

# Speed at a particular given time
INST_SPEEDS <- speeds(buffalo[[1]], FITS[[1]])

head(INST_SPEEDS)

# Add section about distance

```

## Key functions reference

| Function | Package | Purpose | 
|:---|:---|:---|
| `speed()` | `ctmm` | Estimate the average speed of a tracked animal | 
| `speeds()` | `ctmm` | Calculate the instantaneous speed of a tracked animal at any given time |
| `meta()` | `ctmm` | Estimate population-level mean parameters from individual movement models and estimates (meta-analysis) |



