---
title: "applying clustering on movement metrics"
author: "Wenjing Xu 徐雯靓 (wenjing.xu@senckenberg.de)"
date: '`r format(Sys.time(), "%d %B %Y")`'
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: true
      smooth_scroll: true
    number_sections: false
    #theme: united
    #highlight: tango
pdf_document: default
---

```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<a href="https://github.com/Animal-Movements/China_2025.git" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#151513; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>
```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(amt)
library(ggplot2)
```

## 1. Introduction

The goal of this analysis is to identify distinct movement strategies among wildebeest individuals based on movement-derived summary metrics. We apply both hierarchical and k-means clustering to explore individual-level patterns in:
- displacement
- path straightness
- intensity of space use
- minimum convex polygon (MCP) area


The two main clustering methods used in this workshop are:
- **Hierarchical clustering**, which builds a tree-like structure of nested groups without pre-specifying the number of clusters.
- **K-means clustering**, which partitions the data into a pre-defined number of clusters by minimizing within-cluster variance.


Why clustering?
- Clustering provides an unsupervised, data-driven way to group individuals
  based on multivariate movement patterns.
- It is simpler and more interpretable than behavioral segmentation methods (e.g. state‑space models), especially at the individual level.
- Clustering enables exploration of general movement syndromes before applying more complex behavioral modeling.

## 2. Load and Prepare Data

```{r load-data}
load("./Data/wildebeest_data.rdata")

WB <- WB %>%
  as_tibble() %>%
  filter((fixType == "3D" & as.numeric(DOP) < 10) |
         (fixType == "2D" & as.numeric(DOP) < 5))

WB_summary <- WB %>%
  arrange(animal_id, timestamp) %>%
  group_by(animal_id) %>%
  summarise(
    start_time = min(timestamp, na.rm = TRUE),
    end_time = max(timestamp, na.rm = TRUE),
    total_duration_days = as.numeric(difftime(end_time, start_time, units = "days")),
    interval_h = median(as.numeric(diff(timestamp), units = "hours"), na.rm = TRUE),
    n_fixes = n(),
    .groups = "drop"
  )

# some individuals have less than one year amount of data
# to make sure the movement metrics are comparable, we subset the dataset 
# to contain only individuals with at least one year amount of data 
# and subset the data to contain only data between Nov 1st 2010 and Oct 31 2011. 

long_term_ids <- WB_summary %>%
  filter(total_duration_days >= 365) %>%
  pull(animal_id)

WB.one.year <- WB %>%
  filter(animal_id %in% long_term_ids) %>%
  filter(timestamp >= ymd("2010-11-01") & timestamp <= ymd("2011-10-31"))

# then calculate metrics for all individuals in the one year dataset
wb.trk <- make_track(WB.one.year, 
                     # tell the function which column contains the coordinates
                     .x = longitude, .y = latitude, 
                     # tell the function which CRS the coordinates are based on. 
                     crs = 4326, 
                     .t = timestamp,
                     # any additional information you want to attach to the track
                     # such as animal ID and month
                     id = animal_id, sex = sex, temp = temp) 

wb.trk.nested <- wb.trk %>%
  transform_coords(., 32737) %>% 
  nest(data = -"id")

```

## 3. Define and Compute Movement Metrics

```{r metrics-functions}
calc_disp <- function(track_df, time_unit = c("day", "week")) {
  time_unit <- match.arg(time_unit)
  
  track_df %>%
    mutate(period = floor_date(t_, unit = time_unit)) %>%
    arrange(t_) %>%
    group_by(period) %>%
    summarise(
      x_start = first(x_),
      y_start = first(y_),
      x_end   = last(x_),
      y_end   = last(y_),
      .groups = "drop"
    ) %>%
    mutate(
      displacement = sqrt((x_end - x_start)^2 + (y_end - y_start)^2)
    )
}

# calc maximum displacement monthly and annually 
calc_max_disp <- function(track_df, time_unit = c("month", "year")) {
  
  if (time_unit == "year") {
    coords <- track_df %>% select(x_, y_) %>% as.matrix()
    # Get all pairwise distances efficiently
    dists <- dist(coords)
    max_dist <- max(dists, na.rm = TRUE)
    
    tibble(
      max_disp = max_dist
    )
  } else {
    
    df <- track_df %>%
      mutate(period = floor_date(t_, unit = time_unit)) %>%
      group_by(period) %>%
      group_split()
    
    map_dfr(df, function(df_period) {
      n <- nrow(df_period)
      if (n < 2) return(tibble(period = df_period$period[1], max_disp = NA_real_))
      
      coords <- df_period %>% select(x_, y_) %>% as.matrix()
      
      # Get all pairwise distances efficiently
      dists <- dist(coords)
      max_dist <- max(dists, na.rm = TRUE)
      
      tibble(
        max_disp = max_dist
      )
    })
  }
}

# calculate monthly or yearly straightness
calc_straightness <- function(track_df, time_unit = c("month", "year")) {
  time_unit <- match.arg(time_unit)
  
  if (time_unit == "year") {
    # Calculate straightness once over all points
    straight <- straightness(track_df)
    return(tibble(period = "all_year", straightness = straight))
  } else {
    track_df <- track_df %>%
      mutate(period = floor_date(t_, unit = time_unit)) %>%
      arrange(t_)
    
    periods <- unique(track_df$period)
    
    map_dfr(periods, function(p) {
      track_period <- filter(track_df, period == p)
      
      straight <- straightness(track_period)
      
      tibble(period = p, straightness = straight)
    })
  } 
  
}

# calculate monthly or yearly intensity of use 
calc_intensity_use <- function(track_df, time_unit = c("month", "year")) {
  time_unit <- match.arg(time_unit)
  
  if (time_unit == "year") {
    # Calculate intensity_use once over all points
    intensity <- intensity_use(track_df)
    return(tibble(period = "all_year", intensity_use = intensity))
  } else {
    track_df <- track_df %>%
      mutate(period = floor_date(t_, unit = time_unit)) %>%
      arrange(t_)
    
    periods <- unique(track_df$period)
    
    map_dfr(periods, function(p) {
      track_period <- filter(track_df, period == p)
      
      intensity <- intensity_use(track_period)
      
      tibble(period = p, intensity_use = intensity)
    })
  } 
}


# calculate monthly intensity of use 
calc_mcp <- function(track_df, time_unit = c("month", "year")) {
  time_unit <- match.arg(time_unit)
  
  if (time_unit == "year") {
    # Calculate intensity_use once over all points
    mcp95 <- hr_mcp(track_df, levels = 0.95)$mcp$area
    return(tibble(period = "all_year", mcp95 = mcp95))
  } else {
    track_df <- track_df %>%
      mutate(period = floor_date(t_, unit = time_unit)) %>%
      arrange(t_)
    
    periods <- unique(track_df$period)
    
    map_dfr(periods, function(p) {
      track_period <- filter(track_df, period == p)
      
      if (nrow(track_period) < 2) {
        return(tibble(period = p, intensity_use = NA_real_))
      }
      
      mcp95 <- hr_mcp(track_period, levels = 0.95)$mcp$area
      
      tibble(period = p, mcp95 = mcp95)
    })
  }
}
```

With the functions defied, we can now compute all the metrics. 
```{r compute-metrics}
wb_daily_disp <- wb.trk.nested %>%
  mutate(daily_disp = map(data, ~ calc_disp(.x, time_unit = "day"))) %>%
  select(id, daily_disp) %>%
  tidyr::unnest(daily_disp) %>%
  group_by(id) %>%
  summarise(ave_daily_disp = mean(displacement, na.rm = T),
            sd_daily_distp = sd(displacement, na.rm = T))

# For monthly max displacement
wb_monthly_max_disp <- wb.trk.nested %>%
  mutate(monthly_disp = map(data, ~calc_max_disp(.x, time_unit = "month"))) %>%
  select(id, monthly_disp) %>%
  unnest(monthly_disp) %>%
  group_by(id) %>%
  summarise(ave_max_monthly_disp = mean(max_disp, na.rm = T),
            sd_max_monthly_disp = sd(max_disp, na.rm = T))

# For yearly max displacement
wb_yearly_max_disp <- wb.trk.nested %>%
  mutate(yearly_disp = map(data, ~calc_max_disp(.x, time_unit = "year"))) %>%
  select(id, yearly_disp) %>%
  unnest(yearly_disp) %>%
  rename(yearly_max_disp = max_disp)

# For monthly straightness
wb_monthly_straightness <- wb.trk.nested %>%
  mutate(monthly_straightness = map(data, ~ calc_straightness(.x, time_unit = "month"))) %>%
  select(id, monthly_straightness) %>%
  tidyr::unnest(monthly_straightness) %>%
  group_by(id) %>%
  summarise(ave_monthly_str = mean(straightness, na.rm = T),
            sd_monthly_str = sd(straightness, na.rm = T))

wb_yearly_straightness <- wb.trk.nested %>%
  mutate(yearly_straightness = map(data, ~ calc_straightness(.x, time_unit = "year"))) %>%
  select(id, yearly_straightness) %>%
  tidyr::unnest(yearly_straightness) %>%
  select(-period) %>%
  rename(yearly_straightness = straightness)

# monthly intensity of use per animal
wb_monthly_intensity <- wb.trk.nested %>%
  mutate(monthly_intensity = map(data, ~ calc_intensity_use(.x, time_unit = "month"))) %>%
  select(id, monthly_intensity) %>%
  tidyr::unnest(monthly_intensity) %>%
  group_by(id) %>%
  summarise(ave_monthly_intensity = mean(intensity_use, na.rm = T),
            sd_monthly_intensity = sd(intensity_use, na.rm = T))

wb_yearly_intensity <- wb.trk.nested %>%
  mutate(annual_intensity = map(data, ~ calc_intensity_use(.x, time_unit = "year"))) %>%
  select(id, annual_intensity) %>%
  tidyr::unnest(annual_intensity) %>%
  select(-period) %>%
  rename(yearly_intensity = intensity_use)

# monthly MCP of use per animal
wb_monthly_mcp <- wb.trk.nested %>%
  mutate(monthly_mcp = map(data, ~ calc_mcp(.x, time_unit = "month"))) %>%
  select(id, monthly_mcp) %>%
  tidyr::unnest(monthly_mcp) %>%
  group_by(id) %>%
  summarise(ave_monthly_mcp = mean(mcp95, na.rm = T),
            sd_monthly_mcp = sd(mcp95, na.rm = T))

wb_yearly_mcp <- wb.trk.nested %>%
  mutate(annual_mcp = map(data, ~ calc_mcp(.x, time_unit = "year"))) %>%
  select(id, annual_mcp) %>%
  tidyr::unnest(annual_mcp)%>%
  select(-period) %>%
  rename(yearly_mcp = mcp95)
```

Combine all metrics into a single data frame 
```{r data combo}
metrics.list <- list(wb_daily_disp,
                     wb_monthly_max_disp,
                     wb_yearly_max_disp,
                     wb_monthly_straightness,
                     wb_yearly_straightness,
                     wb_monthly_intensity,
                     wb_yearly_intensity,
                     wb_monthly_mcp,
                     wb_yearly_mcp) 
metrics.df <- reduce(metrics.list, ~ left_join(.x, .y, by = "id")) 
```

## 4. Clustering Analysis
First we need to convert the dataframe into a matrix. Because different metrics have different units and ranges, we scale the matrix to ensures equal weight across behavioral axes
```{r clustering}
# Standardize metrics to mean = 0, unit variance
metrics.mat <- scale(as.matrix(metrics.df %>% select(-id))) 
rownames(metrics.mat) <- metrics.df$id
```

```{r HIERARCHICAL CLUSTERING}
# Compute distance matrix and apply clustering method
dist_mat <- dist(metrics.mat)
hc <- hclust(dist_mat, method = "ward.D2") # ward.D2 creates compact, spherical clusters (often fits behavioral mode clustering).
# Visualization
plot(hc, main = "Hierarchical Clustering Dendrogram")
```

```{r k-means}
set.seed(123)# reproducibility
kmeans_result <- kmeans(metrics.mat, centers = 3, nstart = 25)
metrics.df$cluster <- as.factor(kmeans_result$cluster)

# Visualize using PCA projection onto first two principal components
pca <- prcomp(metrics.mat)
pca_df <- data.frame(pca$x[,1:2], id = rownames(metrics.mat), cluster = metrics.df$cluster)

ggplot(pca_df, aes(PC1, PC2, color = cluster, label = id)) +
  geom_point(size = 3) +
  geom_text(vjust = 1.5, hjust = 1.2, size = 3) +
  theme_minimal() +
  labs(title = "K-means Clustering (PCA Projection)", color = "Cluster")
```

|           <span style="color:tomato">**QUESTION: ** From the track maps we created before, can you try to interpret the clustering results? </span>
<br/><br/>


---------------------
Further readings:

+ Abrahms, Briana, Dana P. Seidel, Eric Dougherty, Elliott L. Hazen, Steven J. Bograd, Alan M. Wilson, J. Weldon McNutt et al. "Suite of simple metrics reveals common movement syndromes across vertebrate taxa." Movement ecology 5, no. 1 (2017): 12.

 <center> -----------------     (ﾉ◕ヮ◕)ﾉ*:･ﾟ✧  THE END :･ﾟ･ﾟ     ------------------ </center>
