---
title: 'M4: Accounting for Location Error'
subtitle: "Day 5 — Behavior Analysis Part 2"
author: "Animal Movement Analyses for Ecologists | SYSU Shenzhen, July 2026"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    toc_collapse: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Created: 2026-07-13* | *Last updated: `r Sys.Date()`*

<a href="https://github.com/Animal-Movements/China_2026.git" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#151513; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

---

## Learning Objectives

By the end of this module you will be able to:

1. Fit an error model to calibration data and incorporate the error estimates into the telemetry data.
2. Assess the fit of the error model, by checking the residuals.
3. Include error estimates from the calibrated data into movement analyses.
4. Identify outliers in the data that are consistent or inconsistent with the error model.

**Packages used:** `ctmm`

---

## Background

## Data Preparation

```{r import_data}

# load ctmm library
library(ctmm)

# load turtle data
data(turtle)
# or as.telemetry on the turtle data filename

# turtle datasets
names(turtle)
# first two are calibration data - not turtles

# second look at columns
head(turtle[[1]])
# HDOP: horizontal dilution or precision -- proportional to RMS error (location error w/ location class)
# location class: 3D (or 2D) -- generally doesn't matter (calculated on same scale), but some companies use diff algorithms for 2D vs 3D
## HDOP: want proportionality constant close to 1

```

## Fitting an Error Model to Calibration Data

```{r uere.fit, eval = FALSE}

## Assuming all of the information is good

help("uere.fit")  # user equivalent range error 

```

```{r uere.fit2}

# fit error parameters to calibration data (informs error model)
UERE <- uere.fit(turtle[1:2])
# do not run uere.fit on tracking data (using calibration data, not moving turtles)

# estimated error model parameters
summary(UERE)
## 3D fixes have ~7 at HDOP 1
## 2D fixes have ~ 30 at HDOP 1 (diff scales)

# apply error model to data
uere(turtle) <- UERE

head(turtle$F231)

plot(turtle$F231)  # error circles at each tracked location accounting for calibration data

## If we aren't sure about the error data:
## QUESTION 1: Are the HDOP and location class values informative?
data(turtle)

# make a list to store error models
UERES <- list()

# first attempt: let's use everything
UERES$all <- uere.fit(turtle[1:2])
# do not run uere.fit on tracking data

# summarize error model
summary(UERES$all)

# second attempt: let's drop the location class information
# copy of calibration data
test <- turtle[1:2]
# delete location class column
test[[1]]$class <- NULL
test[[2]]$class <- NULL
uere(test) <- NULL
# store error-model fit (HDOP only)
UERES$HDOP <- uere.fit(test)

# summarize error model
summary(UERES$HDOP)  # removing location class data (only HDOP), ~10.6 m error

# third attempt: let's further drop the HDOP values
# delete HDOP column
test[[1]]$HDOP <- NULL
test[[2]]$HDOP <- NULL
# store error-model fit
UERES$nada <- uere.fit(test)

# summarize error model
summary(UERES$nada)  # removing HDOP as well (large location error estimate)

# compare error-models
summary(UERES)  # AICc selects model w/ all, but all have poor Z^2 values because HDOP values aren't great
# AICc: super-fancy AIC values
# reduced Z-squared statistic (goodness of fit for trends)
# compare to reduced chi-squared statistic (1 = good regression, <1 = suspicious, too good, >1 = bad)

```

```{r uere.fit3}

## QUESTION 2: Are these GPS tags identical? (heterogeneity of GPS tags)

# create a list to store individualized error models
indiv <- list()

# calculate individual UEREs
indiv[[1]] <- uere.fit(turtle[[1]])
indiv[[2]] <- uere.fit(turtle[[2]])

# compare calibration parameters
summary(UERES$all) # joint model
## ~7m error for 3D, ~30m for 2D
summary(indiv[[1]]) # but pretty similar if fit separately
summary(indiv[[2]])

# store with joint models
UERES$indiv <- indiv

# compare to joint models
summary(UERES)
# Don't trust AICc here, because it will likely just pick the more complex model (doesn't actually improve much over indiv)

```

## Checking Error Model Fit

```{r assign_error}

#############
# ERROR CALIBRATION
#############

# calibrate turtle data with best error model
uere(turtle) <- UERES$all  # assign error model to be the error of dataset

# error columns now in data
head(turtle[[1]])
# also now includes a per-time error estimate (error circle)
plot(turtle[[1]]) # see error in indiv turtle movement (most of the variation is explained by error)

```

```{r residuals}

#############
# ERROR-MODEL RESIDUALS
#############

# calculate residuals of calibration data w.r.t best error model
RES <- lapply(turtle[1:2],residuals)
# plot residuals
plot(RES)  # residuals in red (st.dev of null model, w/ 95% CI)
## bad error models would give heavy tail distr

# calculate residuals of calibration data w.r.t. worst error model
uere(test) <- UERES$nada  # homoskedastic error model

RES2 <- lapply(test,residuals)
plot(RES2)

```

## Error-Informed Movement Analysis

```{r outlie}

# turtle data again
names(turtle)

# take female turtle 231
DATA <- turtle$F231

# plot data
plot(DATA)

help('outlie')  # function to help pick out outliers (using error-informed statistics)

# look for outliers
OUT <- outlie(DATA)  # side effect plot (blue segment = highest speed, used to determine outliers)

plot(OUT)
# you may get other useful information here
head(OUT)

# biological decision to reject all speeds > 5 cm/s for this species (wood turtle)
MAX <- 5 %#% 'cm/s'
?'%#%' # convenient units function
MAX # 5 cm/s in SI units
# alternatively
MAX <- 0.05 # in meters/second

# good location estimates
GOOD <- OUT$speed < 0.05 # biological threshold for this species (wood turtle)
# threshold speed for turtle (can remove data above that speed, because those are unreasonable outliers)
# always check errors and outliers just in case because sometimes there's unclear delineation of threshold
## or GPS signals could bounce around, which affects path lengths (gives larger errors than expected)

# take only good location estimates
DATA <- DATA[GOOD,]

# re-check
plot(DATA)
OUT <- outlie(DATA)
# Note now all the points are similarly distanced from median and multiple quickest movement speeds
## data looks more reasonable
plot(OUT)

```

```{r ctmm.guess_error}

# create guesstimate interactively
ctmm.guess(DATA)
# * check the error box (turns error on, not on by default because it's slow)

# create guesstimate non-interactively
GUESS <- ctmm.guess(DATA,CTMM=ctmm(error=TRUE),interactive=FALSE)
# new argument CTMM, which can contain extra parameters,
# here error=TRUE

# fit models
FITS <- ctmm.select(DATA,GUESS,verbose=TRUE,trace=3,cores=-1)
# verbose=TRUE returns all candidate models
# I've already run this code for you
# save(FITS,file="Data/turtle.rda")
load("Data/turtle.rda")

# look at all models
summary(FITS)
## location error model turned on

# look at best model
summary(FITS[[1]])
## updated error parameter estimates (still similar to what was fed in)

# compare to prior model
summary(uere(DATA))

# compare to movement model without error model 
GUESS <- FITS[[1]]
GUESS$error <- FALSE
FIT.NE <- ctmm.fit(DATA,GUESS,trace=2)  # fits quickly w/out error model turned on

summary(FITS[[1]])$CI  # with error model ( also allows comparison btwn two datasets/studies using diff GPS tags)
summary(FIT.NE)$CI  # without error model (larger HR estimate due to all variation being explained by movement)
## bigger movement rates too, some jittering of data should be explained by location error, 
## but w/out error model, it's all being explained by animal movement

## Smoothing data for other packages (not ctmm)

help('predict', package="ctmm")

SMOOTH <- predict(DATA,FITS[[1]])  # smoothed data to update, removes some error because it feeds in the movement model

plot(DATA)
plot(SMOOTH)  # updated estimated of turtle locations (uses error ellipses, not circles)

SIM <- simulate(DATA,FITS[[1]])  # simulate based on movement + error model
plot(SIM)

```

## No Calibration Data? Supply a Prior

```{r prior}

## IF YOU DIDN'T HAVE CALIBRATION DATA, SUPPLY A PRIOR

# load un-calibrated datas
data(turtle)

# will need to match the class structure (2D,3D here)
summary(uere(turtle))  # guesses for error (10m)

# supply point estimates (assign numbers to error)
# 20-meter 2D error at HDOP=1
# 10-meter 3D error at HDOP=1
uere(turtle) <- c(20,10)

# extract calibration object
PRIOR <- uere(turtle)
# the default uncertainty when assigning numerical error is zero
summary(PRIOR)
PRIOR$DOF

# set DOF for wide credible intervals
PRIOR$DOF[] <- 2  # Bayesian: error is worth 2 data points
summary(PRIOR)  # data calibrated w/ prior (much more uncertainty due to few data points)

# assign prior to data
uere(turtle) <- PRIOR

# automated guesstimate for calibrated data
GUESS <- ctmm.guess(turtle[[3]],CTMM=ctmm(error=TRUE),interactive=FALSE)
FIT.PRIOR <- ctmm.select(turtle[[3]],GUESS,trace=3,cores=-1)
# this will take a while, but comes out consistent
# save(FIT.PRIOR,file="Day5_BehaviourAnalysis_PT2/Data/turtle-prior.rda")
load("Day5_BehaviourAnalysis_PT2/Data/turtle-prior.rda")

summary(FIT.PRIOR)  # using prior and updating it

# compare update to prior
summary(PRIOR)
## further updated due to larger DOF (bigger sample size)
## but be cautious (better than nothing but not a substitute for actual calibration data)
## may just get same prior back regardless of updates (best you'll get, because this is a last resort to adding an error model)

```

---

## Key functions reference

| Function | Package | Purpose | 
|:---|:---|:---|
| `uere.fit()` | `ctmm` | Fit error parameters to GPS calibration data to estimate the root-mean-square User Equivalent Range Error (UERE) of a GPS device | 
| `uere()` or `uere()<-` | `ctmm` | Assign the UERE error model to telemetry data |
| `ctmm()` | `ctmm` | Fit a hypothetical movement model by specifying the model parameters; only used in this case to specify `error = TRUE` for the error model to be incorporated in the `ctmm.guess` function |
| `outlie()` | `ctmm` | Detect outliers in telemetry data, informed by an error model, by highlighting potential speed and distance outlier (or anomalous) points |

---

_Next: **Day 6 — Habitat Selection** — We will examine the habitat preferences (or avoidances) of individuals through habitat selection analysis with resource selection functions (RSFs) and step-selection functions (SSFs)._




