---
title: 'M3: Hidden Markov Modeling for Accelerometer Data'
subtitle: "Day 5 — Behavior Analysis Part 2"
author: "Animal Movement Analyses for Ecologists | SYSU Shenzhen, July 2026"
date: '`r format(Sys.time(), "%d %B %Y")`'
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: true
      smooth_scroll: true
    number_sections: false
pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<a href="https://github.com/Animal-Movements/China_2026.git" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#151513; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

---

# Learning Objectives
By the end of this module you will be able to:

1. Fit a HMM to accelerometer data
2. Become comfortable with the Hmm workflow
3. Visualize output

**Packages used:** `momentuHMM`, `lubridate`, `tidyverse`

# Introduction
  >Many thanks to Dr. Dongmin Kim for assistance in developing this material

We've now learned the basics of fitting an HMM. We'll now take this one step further, incorporating data from accelerometers fitted on maned wolves (*Chrysocyon brachyurus*) in Parque Nacional da Serra da Canastra in Brazil.In these analyses, we will focus on two individuals (male: ID 32866 - Juba; female: ID 32733 - Pacoca). Juba and Pacoca are a mating pair and Pacoca was pregnant when she was fitted with the accelerometer.  Many thanks to Dr. Rosanna N. Moraes for making these data available for use. 

```{r load-pacakages, message=FALSE, warning=FALSE, echo=TRUE}
# Remove items from memory/clean your workspace
rm(list=ls())

# You may need to install these packages first
#install.packages('momentuHMM', 'lubridate', 'tidyverse')

# Load libraries
library(momentuHMM)
library(lubridate)
library(tidyverse)
```

# 1. Load Data
The accelerometer data used in this module are saved in a proprietary format (i.e., .adf) from the  manufacterer (Vectronics Aerospace). In order to load the data into [R](https://cran.r-project.org/), I first used the manufacturer's software to convert them to a R-readable format (i.e., .csv). 

For the purposes of this demonstration, we take the summarized activity data from the raw accelerometer from two individuals: ID 32866-Juba and ID 32733-Pacoca. Accelerometer data is different than GPS data, with activity data provided at much finer temporal scales. 

```{r load-data, message=FALSE, warning=FALSE, echo=TRUE}
# Load data 

## ID: 32733 - pacoca
act_32733 <- read.csv("data/ACT_Collar32733_20260401143616.csv")

## ID: 32866 - juba
act_32866 <- read.csv("data/ACT_Collar32866_20260401143833.csv")
```

# 2. Clean Timestamps 
Note that the date and time columns (UTC_Date and UTC_Time, respectively) are organized as separately.  Because these columns are "character" columns, we need to join them together and then convert them to a date/time field.

```{r time-conversion, message=FALSE, warning=FALSE, echo=TRUE}
## currently, the data contains several date/time columns -- date and time need to combine into one column first 
act_32733$datetime_str <- paste(act_32733$UTC_Date, act_32733$UTC_Time)
act_32866$datetime_str <- paste(act_32866$UTC_Date, act_32866$UTC_Time)

## convert to POSIXct 
act_32733$time <- lubridate::mdy_hms(act_32733$datetime_str, tz = "UTC")
act_32866$time <- lubridate::mdy_hms(act_32866$datetime_str, tz = "UTC")
```

The base time interval between observations was 5 minutes (300 secs) in this study. We need to make sure to check if there were missing observations, resulting in longer intervals that need to be corrected.  

```{r diff-time, message=FALSE, warning=FALSE, echo=TRUE}
# Table of time intervals in data - can apply to both individuals
plot(table(diff(act_32866$time)), xlim = c(0, 600),
     xlab = "time interval (sec)", ylab = "count")
```

Although our data do not have issues with temporal irregularity, it is always important to check and deal with temporal irregularity before fitting a HMM.  You now know how to do this, which would require a few additional steps to fill in the data with NAs so that the temporal interval was consistent.

## 2.1 Calculating ODBA
Unlike GPS data, activity data do not include location information such as coordinates (e.g., Easting/Northing or latitude/longitude). Instead, activity data capture movement using proxy variables. In our data, these activity data are labeled as 'ActivityX', 'ActivityY', and 'ActivityZ', measuring motion along three dimensions (X, Y, and Z - surge, sway, and heave). These values are recorded at summarized at regular intervals (e.g., every 5 minutes) and reflect how the animal is moving rather than where it is.

These three measurements are often combined into a single summary metric called *Overall Dynamic Body Acceleration* or *ODBA*, representing the overall intensity of movement. ODBA is commonly used as an indicator of activity or energy expenditure. Let's calculate ODBA for our data.

```{r ODBA, message=FALSE, warning=FALSE, echo=TRUE}
## use a single summary activity variable - ODBA
act_32733$ODBA <- rowSums(act_32733[, c("ActivityX", "ActivityY", "ActivityZ")], na.rm= TRUE)
act_32866$ODBA <- rowSums(act_32866[, c("ActivityX", "ActivityY", "ActivityZ")], na.rm = TRUE)
```

## 2.2 Further Data Wrangling 
Let's select several columns of interest for the analysis and combine both animals into a single dataframe.

```{r final-data, message=FALSE, warning=FALSE, echo=TRUE}
# select the columns of interest
pacoca <- act_32733 %>% 
  select(ID = CollarID, 
         ODBA, 
         time, 
         temp = Temp...C.)

juba <- act_32866 %>% 
  select(ID = CollarID, 
         ODBA, 
         time, 
         temp = Temp...C.)

# combine the datasets
data <- rbind(pacoca, juba)

# Summarize
summary(data)
```

# 3. The MomentuHMM Workflow
The data are now ready to be used in [momentuHMM](https://cran.r-project.org/web/packages/momentuHMM/index.html), following a similar set of steps that we are now familiar with, inclusive of the covariate temperature.  

## 3.1 Data preparation
Note that we do not handle the location data like the GPS example. With GPS data, we derive step lengths and turning angles from the locations. Here, we set the 'coordNames' argument to NULL within the function 'prepData'. 

```{r prep-data1, message=FALSE, warning=FALSE, echo=TRUE}
# Prepare data for HMM 
data_hmm <- prepData(data, coordNames = NULL, covNames = "temp")
```

## 3.2 Model Fitting 
[MomentuHMM](https://cran.r-project.org/web/packages/momentuHMM/index.html) is quite flexible and can handle multiple datastreams, inclusive of GPS data for coarse-scale movements. Here, we will focus intently on the activity data only, using the function `fitHMM()` to fit models like we used previously.

We are interested in two states, State 1 or *'Low Activity'* and State 2 or *'High Activity'*.  To get started, we need to specify two things: 

1. Observation distributions - Each observed variable needs a probability distribution to define each activity state. ODBA is continuous and strictly positive (except for exact zeros), allowing us to once again use the gamma distribution. Since many zeros exist in our data, we need to use a zero-inflated gamma distribution and account for zero mass as we did in our previous analysis. 

2. Initial parameter values - The `fitHMM()` function uses numerical optimization to find maximum likelihood estimates. The optimization requires a starting point. For a gamma distribution with zero-inflation, each state requires three parameters. As before, these are: (1) mean ($\bar{x}$) of the gamma, (2) standard deviation ($\sigma{x}$) of the gamma, and (3) probability of exact zero (i.e., zero-mass) for each state.

Please once again refer to the [moveHMM](https://cran.r-project.org/web/packages/moveHMM/index.html) guide on selecting initial parameter values, as this can be very tricky: [A short guide to choosing initial parameter values for the estimation in moveHMM](https://cran.r-project.org/web/packages/moveHMM/vignettes/moveHMM-starting-values.pdf).

```{r fit-hmm1, message=FALSE, warning=FALSE, echo=TRUE}
# Observation distributions (ODBA)
dist <- list(ODBA = "gamma")

# Initial parameters
# Use simple rules-of-thumb based on the data:
# - Low activity state (state 1): roughly half the mean ODBA
# - High activity state (state 2): roughly 1.5 times the mean ODBA
# The 1e-2 ensures strictly positive values.  Not necessary here, but might save you some problems.
mu1 <- max(mean(data$ODBA) * 0.5, 1e-2)  # state 1: low activity 
mu2 <- max(mean(data$ODBA) * 1.5, 1e-2)  # state 2: high activity

# Use the standard deviation of the positive observations as a guide.
# Ensure strictly positive values (1e-2) to satisfy momentuHMM bounds.
sd1 <- max(sd(data$ODBA) * 0.5, 1e-2)
sd2 <- max(sd(data$ODBA) * 1.5, 1e-2)

# Zero-inflation: probability of observing exact zeros
# Can estimate from data, or use small non-zero values as starting points -- Must be strictly < 1
zm1 <- 0.01
zm2 <- 0.01

# Combine into Par0
# The order is important: c(mu1, mu2, sd1, sd2, zm1, zm2)
Par0_2s <- list(
  ODBA = c(mu1, mu2, sd1, sd2, zm1, zm2)
)

# Fit a 2-state HMM
hmm1 <- fitHMM(data_hmm, 
               nbStates = 2, 
               dist = dist, 
               Par0 = Par0_2s)

# Print parameter estimates 
hmm1
```

## 3.3 Model Visualization 
Similar to previous exercises, we see summary statistics and regression coefficients for model output. We can also plot the estimated distributions to visualize the findings.

```{r look-hmm1, fig.keep = 1:4}
# Plot estimated distributions and state-coloured tracks
plot(hmm1, ask = FALSE)
```

We can also estimate the most likely state sequence for a fitted model with the 'viterbi' algorithm. 

```{r viterbi, message=FALSE, warning=FALSE, echo=TRUE}
# get most likely sequence of states 
head(viterbi(hmm1))

# save most likely state sequence from 2-state model
data_hmm$state_2st <- factor(viterbi(hmm1))

# plot OBDA over time, colored by states
ggplot(data_hmm, aes(x = time, y = ODBA, col = state_2st, group = ID))+
  geom_point(size = 0.5, alpha = 0.2)+
  facet_wrap(~ID)
```

## 3.4 Model Formulation: Covariates
Now, following the same structure as previously, we will include the temperature measurements collected by the tag at each time of observation. Here, we aim to answer if temperature impacts maned wolf activity patterns. To do so, we'll estimate the effect of temperature on the transition probabilities. 

```{r fit-hmm2, message=FALSE, warning=FALSE, echo=TRUE}
# Fit 2-state HMM with temperature covariate 
hmm2 <- fitHMM(data_hmm, 
               nbStates = 2, 
               dist = dist, 
               Par0 = Par0_2s, 
               formula = ~ temp)

# show summary of model 2 
hmm2
```

Visualizae the stationary state probabilities as a function of the temperature covariate. 

```{r look-hmm2}
# plot stationary state probabilities as functions of temperature 
plotStationary(hmm2, 
               plotCI = TRUE)
```

Thes plot suggests that the maned wolves were more likely to be in state 2 (high activity) at low temperatures and in state 1 (low activity) at high temperatures.

## 3.5 Model Comparison 
We then use the function `AIC()` for model selection, comparing the model without the covariate versus model with the temperature covariate.

```{r AIC}
# Compare models using AIC 
AIC(hmm1, 
    hmm2)
```

These results indicate that the model (model 2) with the temperature covariate has more statistical support. 

## 3.6 Model Checking
We can further evaluate whether a model sufficiently represents the key characteristics of the data. One common method of model checking for HMM's is pseudo-residuals. When the model provides a perfect fit, these pseudo-residuals should follow a standard normal distribution. Any departure from this pattern may suggest that the model does not fully capture the data structure.

The function `plotPR()` can be used to genereate three diagnostic plots for each variable:

1. A time series plot 
2. A quantile-quantile (QQ) plot comparing the residuals to a standard normal distribution 
3. An autocorrelation function (ACF) plot. 

Deviations from the diagonal line in the QQ plot signal potential lack of fit, while the ACF plot helps determine whether the model has successfully accounted for temporal dependence in the data.

```{r pseudo-res}
# plot pseudo-residuals for 2-state models 
plotPR(hmm1)
plotPR(hmm2)
```

Goodness-of-fit is slightly better in model 2. However, the model still needs to be tuned to capture more detailed features of the data. This is a good example of how it is important to do model checking before finalizing the model.  

---