---
title: "M2: Data Cleaning & Quality Control"
subtitle: "Day 1 — Introduction to Movement Data"
author: "Animal Movement Analyses for Ecologists | SYSU Shenzhen, July 2026"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<a href="https://github.com/Animal-Movements/China_2026.git" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#151513; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>


---

## Learning Objectives

By the end of this module you will be able to:

1. Explain why GPS data quality control must happen before any movement analysis
2. Use track-level metadata (`mt_track_data()`) to subset a `move2` object to a study area and deployment window
3. Apply a GPS-precision filter based on dilution of precision (DOP) and fix type
4. Identify and remove duplicate or invalid fixes
5. Quantify how each QC step changes the size and composition of your dataset

**Packages used:** `move2`, `tidyverse`, `lubridate`, `sf`, `gt`, `mapview`

---

## Ecological context

A GPS fix is an estimate, not a fact. Every location carries some error, and that error is rarely random or harmless: a handful of bad fixes can inflate step lengths into biologically impossible "teleports," bias a home range estimate outward, or manufacture an apparent "fast directed movement" behavioral state that never happened. For wide-ranging grazers like wildebeest moving across open, fenced, and increasingly developed landscapes, these artifacts can directly mislead conservation conclusions — for example, overestimating how often animals cross a fence line, or how large a population's range truly is.

The Movebank study we downloaded in Module 1 also pools data across several wildebeest populations in Kenya (not just Athi-Kaputiei). Before we can say anything about *this* population's movement, we need to: (1) isolate the right subset of animals and time window, and (2) remove fixes that don't meet a minimum precision standard. This module does both, and gives you a reusable template for QC-ing your own tracking data.

---

## 1. Setup

### 1.1 Load packages

```{r packages}
# install.packages(c("move2", "tidyverse", "lubridate", "sf", "gt", "mapview"))

library(move2)       # move2 spatial data class and helpers
library(tidyverse)   # Data wrangling and visualization
library(lubridate)   # Timestamps
library(sf)          # Spatial vector data
library(gt)          # Formatted summary tables
library(mapview)     # Quick interactive maps
```

### 1.2 Load the raw data from Module 1

```{r load}
load("Data/WB_raw.rdata")   # loads WB.mv2, saved at the end of Module 1

cat("Starting point:", n_distinct(mt_track_id(WB.mv2)), "animals,", nrow(WB.mv2), "fixes\n")
```

We will track how these two numbers change after every QC step, so initialize a small log:

```{r qc_log_init}
qc_log <- tibble(
  step      = "0. Raw data (Module 1 output)",
  n_fixes   = nrow(WB.mv2),
  n_animals = n_distinct(mt_track_id(WB.mv2))
)
```

---

## 2. Explore track-level metadata

Some attributes — sex, study site, deployment start/end — describe an *individual*, not a single GPS fix. In `move2`, these live in the track metadata, separate from the event (fix-level) data.

```{r track_meta}
track_meta <- mt_track_data(WB.mv2)
glimpse(track_meta)
```

```{r study_sites}
# How many study sites are mixed into this Movebank study?
track_meta %>%
  count(study_site, name = "n_animals") %>%
  arrange(desc(n_animals))
```

> **Why does this matter?** This Movebank study pools several wildebeest populations across Kenya. If we don't subset to Athi-Kaputiei before computing movement metrics or home ranges later in the workshop, we would be mixing animals from ecologically distinct landscapes into a single "population" — a classic and avoidable analytical error.

A quick map colored by study site makes the problem (and the fix) obvious:

```{r map_by_site}
WB.sf_all <- sf::st_as_sf(as.data.frame(WB.mv2)) %>%
  mutate(study_site = track_meta$study_site[match(individual_local_identifier,
                                                    track_meta$individual_local_identifier)])

WB.sf_all %>%
  slice_sample(n = 3000) %>%
  mapview(zcol = "study_site", layer.name = "Study site", cex = 2, alpha = 0.7)
```

---

## 3. Subset to study area and deployment window

We restrict to (a) the Athi-Kaputiei Plains study site and (b) each animal's actual deployment window (removing any pre-deployment test fixes or post-retrieval fixes).

```{r subset_site_deploy}
# Build per-track lookup vectors from the track metadata
site_lookup <- setNames(as.character(track_meta$study_site),
                         as.character(track_meta$individual_local_identifier))
on_lookup   <- setNames(track_meta$deploy_on_timestamp,
                         as.character(track_meta$individual_local_identifier))
off_lookup  <- setNames(track_meta$deploy_off_timestamp,
                         as.character(track_meta$individual_local_identifier))

WB.mv2 <- WB.mv2 %>%
  filter(
    site_lookup[as.character(individual_local_identifier)] == "Athi-Kaputiei Plains",
    timestamp >= on_lookup[as.character(individual_local_identifier)],
    timestamp <= off_lookup[as.character(individual_local_identifier)]
  )

cat("After filter:", n_distinct(mt_track_id(WB.mv2)), "animals,", nrow(WB.mv2), "fixes\n")
print(table(mt_track_data(WB.mv2)$study_site))   # expect: only Athi-Kaputiei Plains
```

```{r qc_log_1}
qc_log <- qc_log %>%
  add_row(step      = "1. Subset to study site + deployment window",
           n_fixes   = nrow(WB.mv2),
           n_animals = n_distinct(mt_track_id(WB.mv2)))
```

> **Note:** `site_lookup[individual_local_identifier]` works because `individual_local_identifier` is duplicated onto every fix (event-level), while the lookup vector has one entry per animal (track-level). Indexing a named vector by a character vector returns one matched value per row — no join required.

---

## 4. Quality-filter by GPS precision (DOP × fix type)

**Dilution of precision (DOP)** quantifies how much satellite geometry inflates positional error at the moment of a fix — lower is better. A **3D fix** (with an altitude solution) is generally more reliable than a **2D fix** at the same DOP value, so we apply a stricter threshold to 2D fixes.

A common rule of thumb (and the one we use here): keep 3D fixes with DOP < 10, and 2D fixes with DOP < 5. But the actual cut-off is highly dependent on your specific case. Sometimes all fixes can be useful even if we know they have high DOP.  

> **Behind the scenes:** `move2` automatically attaches a measurement unit (via the `units` package) to many sensor columns, based on the official Movebank attribute dictionary. `gps_dop` is dimensionless, but still arrives tagged as a `units` object rather than a plain number — which breaks `ggplot2` histograms unless `library(units)` is attached. Since DOP has no real physical unit, we simply drop the units class below and treat it as an ordinary number.

```{r units_fix}
WB.mv2 <- WB.mv2 %>% mutate(gps_dop = as.numeric(gps_dop))
```

```{r dop_explore}
# How many fixes of each type, and what does the DOP distribution look like?
WB.mv2 %>%
  as_tibble() %>%
  count(gps_fix_type_raw)
```

```{r dop_before}
dop_before <- WB.mv2 %>%
  as_tibble() %>%
  filter(!is.na(gps_dop), gps_fix_type_raw %in% c("2D", "3D"))

ggplot(dop_before, aes(x = gps_dop)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(data = tibble(gps_fix_type_raw = c("2D", "3D"), thresh = c(5, 10)),
             aes(xintercept = thresh), color = "firebrick", linetype = "dashed") +
  facet_wrap(~ gps_fix_type_raw, scales = "free_y") +
  labs(title = "GPS DOP by fix type — before filtering",
       subtitle = "Red line = common threshold (2D: DOP < 5, 3D: DOP < 10)",
       x = "DOP", y = "Count") +
  theme_minimal()
```

```{r dop_filter}
# Note: fixes with a missing DOP value, or a fix type other than "2D"/"3D",
# evaluate to NA here and are dropped by filter() along with the low-precision fixes.
WB.mv2 <- WB.mv2 %>%
  filter(
    (gps_fix_type_raw == "3D" & gps_dop < 10) |
    (gps_fix_type_raw == "2D" & gps_dop < 5)
  )

cat("After DOP/fix-type filter:", n_distinct(mt_track_id(WB.mv2)), "animals,", nrow(WB.mv2), "fixes\n")
```

```{r dop_after}
WB.mv2 %>%
  as_tibble() %>%
  ggplot(aes(x = gps_dop)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white") +
  facet_wrap(~ gps_fix_type_raw, scales = "free_y") +
  labs(title = "GPS DOP by fix type — after filtering",
       x = "DOP", y = "Count") +
  theme_minimal()
```

```{r qc_log_2}
qc_log <- qc_log %>%
  add_row(step      = "2. GPS precision filter (DOP × fix type)",
           n_fixes   = nrow(WB.mv2),
           n_animals = n_distinct(mt_track_id(WB.mv2)))
```

> **Heads up for your own data:** Not every tag manufacturer reports DOP and fix type the same way (or at all). e-obs and some solar tags report a different precision metric or none; some vendors give HDOP/VDOP separately rather than a single DOP. Check your tag manufacturer's documentation for what quality fields are available, and adapt the filter accordingly.

---

## 5. Remove duplicates and invalid fixes

```{r dedup}
n_before_dedup <- nrow(WB.mv2)

WB.mv2 <- WB.mv2 %>%
  # Drop exact duplicate animal-timestamp combinations
  distinct(individual_local_identifier, timestamp, .keep_all = TRUE) %>%
  # Drop missing or (0,0) coordinates
  mutate(.lon = sf::st_coordinates(.)[, 1],
         .lat = sf::st_coordinates(.)[, 2]) %>%
  filter(!is.na(.lon), !is.na(.lat), !(.lon == 0 & .lat == 0)) %>%
  select(-.lon, -.lat) %>%
  # Re-establish track + time order (required by many move2/amt functions downstream)
  arrange(individual_local_identifier, timestamp)

cat("Removed", n_before_dedup - nrow(WB.mv2), "duplicate/invalid fixes\n")
```

```{r qc_log_3}
qc_log <- qc_log %>%
  add_row(step      = "3. Remove duplicates + invalid coordinates",
           n_fixes   = nrow(WB.mv2),
           n_animals = n_distinct(mt_track_id(WB.mv2)))
```

---

## 6. Quantify the cumulative effect of QC

```{r qc_summary}
qc_log <- qc_log %>%
  mutate(pct_of_raw = round(100 * n_fixes / first(n_fixes), 1))

qc_log
```

```{r qc_gt}
qc_log %>%
  gt() %>%
  opt_row_striping() %>%
  tab_header(title = "QC pipeline: effect on dataset size",
             subtitle = "White-bearded wildebeest, Athi-Kaputiei Plains") %>%
  cols_label(step       = "Step",
             n_fixes    = "Fixes remaining",
             n_animals  = "Animals remaining",
             pct_of_raw = "% of raw fixes") %>%
  cols_align(align = "center", columns = c(n_fixes, n_animals, pct_of_raw))
```

> **Discussion:** Which step removed the most data — the site/deployment subset, or the DOP filter? Is that what you expected? In your own dataset, which QC step do you think will matter most, and why?

---

## 7. Confirm the cleaned dataset

```{r quick_map_clean}
WB.sf_clean <- sf::st_as_sf(as.data.frame(WB.mv2))

WB.sf_clean %>%
  slice_sample(n = 2000) %>%
  mapview(zcol = "individual_local_identifier", layer.name = "Individual", cex = 2, alpha = 0.7)
```

All points should now fall within the Athi-Kaputiei Plains, with no obvious outliers. If you see anything unexpected, go back and re-examine the QC steps above before moving on.

---

## 8. Save output for Module 3

```{r save}
save(WB.mv2, file = "Data/WB_clean.rdata")
cat("Saved cleaned WB.mv2 —", n_distinct(mt_track_id(WB.mv2)), "animals,", nrow(WB.mv2), "fixes\n")
```

---

## Practice: Apply to your own data

Now run the same QC workflow on your own dataset.

1. **Inspect your track metadata** with `mt_track_data()`. Does it include the fields you need to subset (study site, deployment dates, sex)? If a field is missing, where else might that information live (a separate reference sheet, metadata you recorded yourself)?
2. **Check what precision metric your tags report.** Look for DOP, HDOP, fix type, or a manufacturer-specific quality field. If your tags don't report precision at all, what other QC steps (speed filters, spatial outlier checks) might substitute?
3. **Run the duplicate/invalid-coordinate check** — this step is universal regardless of tag type.
4. **Build your own QC accounting table.** How much data survives each step? Is that consistent with what you'd expect given your species and study design?

---

## Key functions reference

| Function | Package | Purpose |
|---|---|---|
| `mt_track_data()` | `move2` | Access track-level (per-individual) metadata |
| `mt_track_id()` | `move2` | Extract track ID vector (use with `n_distinct()` to count animals) |
| `filter()` | `dplyr` | Subset rows — works directly on `move2` objects |
| `distinct()` | `dplyr` | Remove duplicate rows by key columns |
| `arrange()` | `dplyr` | Re-sort rows — use after filtering to restore track + time order |
| `st_coordinates()` | `sf` | Extract numeric x/y columns from a spatial object |
| `st_as_sf()` | `sf` | Convert a data frame to an `sf` object |
| `gt()` | `gt` | Build a formatted summary table |
| `mapview()` | `mapview` | Quick interactive map |

---

*Next: **Module 3 — Trajectory Creation & Visualization** — we convert the cleaned data into an `amt` track, check sampling regularity, and visualize trajectories.*
