---
title: "M3: Trajectory Creation & Visualization"
subtitle: "Day 1 — Introduction to Movement Data"
author: "Animal Movement Analyses for Ecologists | SYSU Shenzhen, July 2026"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<a href="https://github.com/Animal-Movements/China_2026.git" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#151513; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

---

## Learning Objectives

By the end of this module you will be able to:

1. Build a multi-individual `amt` track (`track_xyt`) from cleaned `move2` data, and set/reproject its coordinate reference system
2. Construct per-individual trajectories as `sf` `LINESTRING` objects and map them interactively, optionally overlaying a contextual shapefile
3. Animate multiple individuals' movement through time with `moveVis`, and recognize when an animation is — and isn't — the right tool

**Packages used:** `amt`, `sf`, `move2`, `tidyverse`, `lubridate`, `units`, `mapview`, `moveVis`

---

## Ecological context

A cleaned table of GPS fixes is still just a point cloud. Connecting those points into trajectories — ordered, individual-specific paths through space and time — is what lets us see things a point cloud hides: repeated-use corridors, a fence line an animal travels along but never crosses, or a shift in range that coincides with the start of the rains. For the Athi-Kaputiei wildebeest, navigating a landscape of fences, roads, and expanding settlement at Nairobi's edge, this distinction is not academic — whether an animal's path tracks *along* a barrier or threads *through* a gap has direct implications for connectivity and where conservation interventions should focus.

Trajectories are also how this kind of analysis gets communicated beyond a methods paper. A clear interactive map, or a short animation of a herd moving across a season, often does more to engage land-use planners, park managers, or pastoralist communities than a table of movement statistics ever will. Today we focus on exactly that: turning cleaned point data into trajectories with `sf`, and into a quick animation with `moveVis`, for spatial accuracy and communication. We'll set up the `amt` track we need for this, too — but the quantitative movement metrics it produces (step length, turning angle, speed) are tomorrow's topic, on Day 2.

---

## 1. Setup

### 1.1 Load packages

```{r packages}
# install.packages(c("amt", "sf", "move2", "tidyverse", "lubridate", "units", "gt", "mapview", "moveVis"))

library(amt)         # Tracks, resampling, step metrics
library(sf)          # Spatial vector data — points and lines
library(move2)       # move2 helpers (mt_track_id, etc.)
library(tidyverse)   # Data wrangling and visualization
library(lubridate)   # Timestamps and durations
library(units)       # Unit-aware arithmetic (used by moveVis below)
library(mapview)     # Quick interactive maps
```

### 1.2 Load the cleaned data from Module 2

```{r load}
load("Data/WB_clean.rdata")   # loads WB.mv2, saved at the end of Module 2

cat("Starting point:", n_distinct(mt_track_id(WB.mv2)), "animals,", nrow(WB.mv2), "fixes\n")
```

### 1.3 Define coordinate reference systems

These were defined in Modules 1–2 but aren't stored in the `.rdata` file, so we redefine them here.

```{r crs}
latlong_crs <- "EPSG:4326"   # WGS84 geographic — what move2/Movebank gives us
utm_crs     <- "EPSG:32737"  # UTM Zone 37S — metric, appropriate for Athi-Kaputiei, Kenya
```

---

## 2. Building a multi-individual `amt` track

`amt` is built around the `track_xy`/`track_xyt` classes — lightweight tibbles with standardized `x_`, `y_`, `t_` columns plus an `id` column when there's more than one individual. We need to hand it plain columns, not a `move2`/`sf` object, so the first step is extracting coordinates out of `WB.mv2`'s point geometry.

### 2.1 From `move2` geometry to plain x/y/t/id columns

```{r flatten}
WB_df <- WB.mv2 %>%
  mutate(x = sf::st_coordinates(.)[, 1],
         y = sf::st_coordinates(.)[, 2]) %>%
  as_tibble() %>%
  transmute(id = individual_local_identifier,
            x, y,
            t = timestamp) %>%
  arrange(id, t)

head(WB_df)
```

### 2.2 `make_track()`

```{r make_track}
trk_all <- make_track(WB_df, x, y, t, id = id, crs = latlong_crs)

class(trk_all)     # track_xyt, track_xy, tbl_df, tbl, data.frame
has_crs(trk_all)   # TRUE — crs was set at creation
head(trk_all)
```

> **Note:** `trk_all` holds *all* individuals in one tibble — `id` is just a column, not a grouping the class enforces. Several `amt` functions we'll use on Day 2 (e.g., `summarize_sampling_rate()`, `track_resample()`) operate on a single track at a time and need to be applied per individual via `nest()`/`map()`/`unnest()` — the same pattern used throughout `amt`'s own documentation for multi-animal data.

### 2.3 Reproject to a metric CRS

```{r reproject}
trk_all <- transform_coords(trk_all, utm_crs)

st_crs(trk_all)$epsg   # 32737
```

> **Why now?** Step lengths, speeds, and home-range areas all need distances in meters, not degrees of latitude/longitude. `transform_coords()` wraps `sf::st_transform()` and requires the track's CRS to already be set — which is exactly what `make_track(..., crs = latlong_crs)` did above.

> **Looking ahead:** `trk_all` is now a ready-to-use, projected `amt` track for all individuals. On Day 2, we'll pick it back up to check each animal's empirical sampling rate, resample to regular bursts, and compute step length / turning angle as movement metrics. For the rest of today, we focus on visualizing the raw trajectories themselves.

---

## 3. Trajectories as spatial lines (`sf`)

`amt` steps are great for movement *metrics*, but for mapping a trajectory as a continuous path, plain `sf` `LINESTRING`s are simpler and more directly compatible with other spatial layers (boundaries, land cover, etc.).

### 3.1 Points → projected `sf` → `LINESTRING`

```{r sf_lines}
WB.sf <- sf::st_as_sf(as.data.frame(WB.mv2)) %>%
  sf::st_transform(utm_crs)

WB.lines <- WB.sf %>%
  group_by(individual_local_identifier) %>%
  dplyr::summarise(do_union = FALSE) %>%
  sf::st_cast("LINESTRING")

WB.lines
```

> **Note:** `group_by(individual_local_identifier) %>% summarise(do_union = FALSE)` collapses each animal's points, *in row order*, into one connected line; `st_cast("LINESTRING")` then sets the geometry type. `do_union = FALSE` matters — `st_union()`'s default behavior can reorder points, which would scramble the trajectory into a meaningless path. This is also why we re-sorted by `id` + `timestamp` back in Module 2.

### 3.2 Interactive trajectory map

```{r sf_static_map}
ggplot() +
  geom_sf(data = WB.lines, aes(color = individual_local_identifier), linewidth = 0.4) +
  scale_color_viridis_d(name = "Individual") +
  labs(title = paste("Wildebeest trajectories — Athi-Kaputiei Plains (n =",
                      n_distinct(WB.lines$individual_local_identifier), ")")) +
  theme_minimal()
```

```{r sf_interactive_map}
mapview(WB.lines, zcol = "individual_local_identifier", layer.name = "Trajectories", lwd = 2) +
  mapview(WB.sf %>% slice_sample(n = 2000), zcol = "individual_local_identifier",
          layer.name = "Fixes", cex = 1.5, alpha = 0.6)
```

### 3.3 (Optional) Overlay a shapefile

Any polygon shapefile — a protected-area boundary, a fence network, a land-cover layer — can be laid under or over your trajectories the same way. We use a protected-area boundary purely as one example; this repository doesn't ship that particular shapefile, but if you have one for your own study system, the pattern from last year's materials applies directly:

```{r nnp_overlay, eval=FALSE}
# Read in your boundary (reproject to match your trajectories' CRS):
# boundary <- sf::st_read("Data/your_boundary.shp", quiet = TRUE) %>%
#   sf::st_transform(utm_crs)
#
# mapview(WB.lines, zcol = "individual_local_identifier", layer.name = "Trajectories") +
#   mapview(boundary, col.regions = "green", alpha.regions = 0.3, layer.name = "Reserve boundary")
```

---

## 4. Animating movement through time (`moveVis`)

### 4.1 Why animate?

An animation makes spatial *and* temporal pattern visible at once in a way a static map cannot — a herd moving together, an individual peeling off, a synchronized response to rainfall. It's a communication tool, not an analysis: an animation doesn't replace the quantitative step-length/turning-angle metrics we'll compute on Day 2, but it's often the fastest way to get a non-specialist audience to see the same pattern those metrics describe. `moveVis` (Schwalb-Willmann et al. 2020, *Methods in Ecology and Evolution* 11:664–669, doi:10.1111/2041-210X.13374) builds these animations directly from `move2` objects.

### 4.2 Subsetting for a tractable demo

```{r movevis_subset}
# if you have not installed moveVis
# Install the latest tested development builds of moveVis from r-universe:
# install.packages("moveVis", repos = '16eagle.r-universe.dev') 
# OR install the latest tested development version of moveVis from source via GitHub:
# remotes::install_github("16EAGLE/moveVis") 
library(moveVis)

# moveVis works directly on cleaned move2 objects -- no need to rebuild an amt
# track for this. For a *quick*, in-class demo we restrict to a few
# individuals and a short time window: animating the full population over the
# full deployment can mean thousands of basemap tiles and frames, which is
# impractical to render live.
demo_ids <- unique(as.character(WB.mv2$individual_local_identifier))[1:3]

demo_start <- min(WB.mv2$timestamp)
demo_end   <- demo_start + weeks(2)

WB_demo <- WB.mv2 %>%
  filter(as.character(individual_local_identifier) %in% demo_ids,
         timestamp >= demo_start, timestamp <= demo_end)

cat(n_distinct(mt_track_id(WB_demo)), "animals,", nrow(WB_demo), "fixes in the demo window\n")
```

### 4.3 Align, build frames, and preview

> **Heads up:** the chunks below need a live internet connection (they download OpenStreetMap basemap tiles) and can take a few minutes even for this small a subset. They're marked `eval=FALSE` here so this module knits quickly and reliably — run them interactively in your own session.

```{r movevis_align, eval=FALSE}
# align_move synchronizes all individuals onto a common, regular time grid.
# A coarser resolution here keeps the frame count (and tile downloads)
# manageable for a live demo -- use a finer res for your own analyses.
m <- align_move(WB_demo, res = units::set_units(6, "hours"))

frames <- frames_spatial(
  m,
  path_colours = viridisLite::viridis(n_distinct(mt_track_id(m))),
  map_service  = "osm", map_type = "topographic", alpha = 0.5
) %>%
  add_labels(x = "Longitude", y = "Latitude") %>%
  add_northarrow() %>%
  add_scalebar() %>%
  add_timestamps(type = "label") %>%
  add_progress()

frames[[10]]   # preview one frame partway through the sequence
```

### 4.4 Render the animation

```{r movevis_render, eval=FALSE}
# Requires a rendering backend:
# install.packages(c("magick", "gifski"))
animate_frames(frames, out_file = "wildebeest_movement.gif")
```

---

## Practice: Apply to your own data

Now build and visualize trajectories for your own dataset.

1. **Build an `amt` track** with `make_track()`. Double check `has_crs()` returns `TRUE` before going further — a missing CRS will silently break `transform_coords()` and any distance-based metric we compute from it on Day 2.
2. **Build `LINESTRING` trajectories** for 2–3 individuals and map them. Do the paths match what you'd expect given your species' ecology and your study area's landscape features?
3. **If you have a relevant shapefile** (a boundary, fence line, land-cover layer), overlay it on your trajectory map as in Section 3.3.
4. **(Stretch) Animate a short window** with `moveVis` for a few individuals. What does the animation reveal that the static trajectory map doesn't?

---

## Key functions reference

| Function | Package | Purpose |
|---|---|---|
| `make_track()` | `amt` | Build a `track_xy`/`track_xyt` object from x/y(/t) columns |
| `has_crs()` | `amt` | Check whether a track's CRS is set |
| `transform_coords()` | `amt` | Reproject a track (wraps `sf::st_transform()`) |
| `st_cast()` | `sf` | Convert geometry type (e.g., `POINT` → `LINESTRING`) |
| `st_transform()` | `sf` | Reproject spatial data to a new CRS |
| `mapview()` | `mapview` | Quick interactive map |
| `align_move()` | `moveVis` | Synchronize multiple tracks onto a common time grid |
| `frames_spatial()` | `moveVis` | Build one animation frame per timestep |
| `animate_frames()` | `moveVis` | Render frames to a GIF/video |

---

*Next: **Day 2 — Movement Metrics** — having built and visualized individual trajectories, we pick `trk_all` back up to check each animal's sampling rate, resample to regular bursts, and quantify movement itself: speed, turning angles, and net displacement, as a foundation for behavioral segmentation.*
