---
title: 'M2-a: Occurrence vs. Range Distributions'
author: "Animal Movement Analyses for Ecologists | SYSU Shenzhen, July 2026"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    toc_collapse: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Created: 2026-06-30* | *Last updated: `r Sys.Date()`* 

<a href="https://github.com/Animal-Movements/China_2026.git" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#151513; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

---

## Learning Objectives

By the end of this module you will be able to:

1. Calculate and visualize an occurrence distribution from tracking data.
2. Calculate and visualize a range distribution from tracking data.
3. Understand the difference between occurrence and range distributions, and why home ranges can only be robustly estimated using range distributions.

**Packages used:** `ctmm`

---

## Background

[INSERT BACKGROUND ON RANGE VS OCCURRENCE]

https://doi.org/10.1101/2022.09.29.509951

## Data Preparation

Let's start by going through the standard data preparation workflow for `ctmm`. Here, we demonstrate only on the first buffalo, Cilla. We have already run through the model selection for Cilla, so let's just load in the saved results (i.e., the best fit model). We will use the OUF anisotropic model that best fits Cilla to estimate her range distribution, and use the Brownian motion models to estimate her occurrence distribution.

```{r import_data}

# Load ctmm package
library(ctmm)

# Import data
data(buffalo)
projection(buffalo) <- median(buffalo)  # center projection on geometric median

# Let's focus on the first buffalo Cilla
DATA <- buffalo$Cilla

# Load her model selection results
load("Data/cilla.rda")

```

## Utilization Distributions

[INSERT INTRO ABOUT UTILIZATION DISTRIBUTIONS]

> What is the **occurrence distribution**?
>
> *Answer:* Given a random time *in the sampling period*, where was the animal? 
>
> This is where the animal likely was during the time that it was tracked (in the past), with some uncertainty in that estimate. The space mapped from an occurrence distribution represents **uncertainty** in where the animal has occurred during the sampling period.

> What is the **range distribution**?
>
> *Answer:* At some time in the future/past, *under the same behaviors*, where will the animal be? 
> 
> This is the long-term space use *for continuing behaviors*. The space mapped from a range distribution represents the areas we predict the animal uses throughout the range resident portion of its life -- the animal's **home range**.

### Occurrence Distribution

```{r occurrence}

# Let's store the fitted movement models in a list
FITS <- list("OUF anisotropic" = FIT)  # selected OUF anisotropic model

# Include Brownian motion models

# Isotropic Brownian motion
FITS[["BM"]] <- ctmm.fit(DATA, ctmm(tau = Inf, isotropic = TRUE))  

# Anisotropic Brownian motion
## Note: this one is not as commonly used, but let's throw it in
FITS[["BM anisotropic"]] <- ctmm.fit(DATA, ctmm(tau = Inf))
# square distance moving away proportional to time taken
## fractal/infinite movement path (most appropriate for coarser data)
## doesn't stay w/in HR (diffuses forever), no more appropriate if not necessarily range resident

# you can't compare stationary (IID,OU,OUF) and conditionally stationary (BM,IOU) models with likelihood
summary(FITS)
# but you can compare within
summary(FITS[c("BM", "BM anisotropic")])

SVF <- variogram(DATA,CI="Gauss")

# again, the selected model looks okay
plot(SVF, FITS[[1]])
plot(SVF, FITS[[1]], frac = 0.01)

# the Brownian motion model looks...
plot(SVF, FITS$BM, frac = 0.01)
# why? zoom in (doesn't asymptote, keeps diffusing infinitely, and no initial quadratic (not ballistic))

# conventional (non-dynamic) Brownian bridge (not the selected model)
BB <- occurrence(DATA, FITS$BM)

# occurrence distribution - using the selected model
OD <- occurrence(DATA, FITS[[1]])  # interpolating from distributions of predictions of buffalo occurrence (average distributions tog to get occurrence distribution)
# predicts well for finely sampled data (negatively biased, biases too small for HR)
# still better than the BM model

# Let's visualize the two occurrence distributions
par(mfrow = c(1,2))  # arrange plots into a 1x2 grid (row of 2)
EXT <- extent(list(DATA, BB, OD))  # make extent the largest of the diff distributions

# Plot BM occurrence distribution (BB)
## Assumes buffalo moves randomly btwn any 2 points (biases small, only good for super coarse data)
plot(BB, col.level = NA, ext = EXT, main = "BM Krige (BB)")  

# Plot OUF occurrence distribution
## Would estimate buffalo to be somewhere along the track (especially for finely sampled data)
plot(OD, col.level = NA, ext = EXT, main = "OUF Krige")  

```

### Range Distribution

```{r range, fig.height = 8, fig.width = 8}

# range distribution - using the selected model
RD <- akde(DATA, FITS[[1]])  # HR

# Let's visually compare the data and 3 distributions
par(mfrow = c(2,2))  # arrange plots into a 2x2 grid
EXT <- extent(list(DATA, OD, RD))  # make extent the largest of the diff distributions

# Plot the raw telemetry data
plot(DATA, ext = EXT, main = "Raw Location Points")

# Plot BM occurrence distribution (BB)
## Assumes buffalo moves randomly btwn any 2 points (biases small, only good for super coarse data)
plot(BB, col.level = NA, ext = EXT, main = "BM Krige (BB)")  

# plot OUF occurrence distribution
## Would estimate buffalo to be somewhere along the track (especially for finely sampled data)
plot(OD, col.level = NA, ext = EXT, main = "OUF Krige")  

# Plot OUF range distribution
## Fuzzy and large due to small effective sample size
plot(RD, col.level = NA, col.grid = NA, ext = EXT, main = "OUF AKDE")  

```

```{r range2}

# Overlay them (`add = TRUE` plots on top of previous)
par(mfrow = c(1,1))  # return to single plot grid
plot(RD, col.level = NA, col.grid = NA, ext = EXT, col.UD = "blue")  # OUF range
plot(OD, col.level = NA, ext = EXT, col.UD = "green3", add = TRUE)  # OUF occurrence
plot(BB, col.level = NA, ext = EXT, col.UD = "red", add = TRUE)  # BM occurrence (BB)
title("Overlayed Distributions")

```

## Sampling Dependency

### Sampling Interval

```{r sampling_freq}

# Impact of COARSENING the data
SUB <- DATA

# Remove every other time
SUB <- SUB[as.logical(1:nrow(SUB)%%2),]  # 2-hr sampling, 4-hr, etc.: as it coarsens, they will begin to look similar and closer to normal distr. (but they aren't the same)

RD <- akde(SUB, FITS[[1]])
OD <- occurrence(SUB, FITS[[1]])

par(mfrow = c(1,2))

plot(RD, col.level = NA, col.grid = NA, ext = EXT)  # range distr doesn't really change as data coarsens
title("Range distribution")
plot(OD, col.level = NA, ext = EXT)  # occurrence distr. becomes increasingly uncertain as data coarsens (sampled at larger time intervals)
title("Occurrence distribution")

```

```{r sampling_freq2}

# Remove every other time
SUB <- SUB[as.logical(1:nrow(SUB)%%2),]  # 2-hr sampling, 4-hr, etc.: as it coarsens, they will begin to look similar and closer to normal distr. (but they aren't the same)

RD <- akde(SUB, FITS[[1]])
OD <- occurrence(SUB, FITS[[1]])

par(mfrow = c(1,2))

plot(RD, col.level = NA, col.grid = NA, ext = EXT)  # range distr doesn't really change as data coarsens
title("Range distribution")
plot(OD, col.level = NA, ext = EXT)  # occurrence distr. becomes increasingly uncertain as data coarsens (sampled at larger time intervals)
title("Occurrence distribution")

```

```{r sampling_freq3}

# Remove every other time
SUB <- SUB[as.logical(1:nrow(SUB)%%2),]  # 2-hr sampling, 4-hr, etc.: as it coarsens, they will begin to look similar and closer to normal distr. (but they aren't the same)

RD <- akde(SUB, FITS[[1]])
OD <- occurrence(SUB, FITS[[1]])

par(mfrow = c(1,2))

plot(RD, col.level = NA, col.grid = NA, ext = EXT)  # range distr doesn't really change as data coarsens
title("Range distribution")
plot(OD, col.level = NA, ext = EXT)  # occurrence distr. becomes increasingly uncertain as data coarsens (sampled at larger time intervals)
title("Occurrence distribution")

```

```{r sampling_freq4}

# Remove every other time
SUB <- SUB[as.logical(1:nrow(SUB)%%2),]  # 2-hr sampling, 4-hr, etc.: as it coarsens, they will begin to look similar and closer to normal distr. (but they aren't the same)

RD <- akde(SUB, FITS[[1]])
OD <- occurrence(SUB, FITS[[1]])

par(mfrow = c(1,2))

plot(RD, col.level = NA, col.grid = NA, ext = EXT)  # range distr doesn't really change as data coarsens
title("Range distribution")
plot(OD, col.level = NA, ext = EXT)  # occurrence distr. becomes increasingly uncertain as data coarsens (sampled at larger time intervals)
title("Occurrence distribution")

# repeat the above until they look similar
# how much data when they look similar?
nrow(DATA)
nrow(SUB)

```

### Sampling Duration

```{r sampling_duration}

# Impact of truncating the data (shorten sampling duration)
SUB <- DATA

# Remove the second half of the data
SUB <- SUB[1:round(nrow(SUB)/2),]  # cutting data in half

RD <- akde(SUB, FITS[[1]])
OD <- occurrence(SUB, FITS[[1]])

par(mfrow = c(1,2))

plot(RD, col.level = NA, col.grid = NA, ext = EXT)  # range distr doesn't change much but will become less detailed (until closer to Gaussian w/ high uncertainty)
title("Range distribution")
plot(OD, col.level = NA, ext = EXT)  # will lose more and more of the distribution until it's a single point
title("Occurrence distribution")

```

```{r sampling_duration2}

# Remove the second half of the data
SUB <- SUB[1:round(nrow(SUB)/2),]  # cutting data in half

RD <- akde(SUB, FITS[[1]])
OD <- occurrence(SUB, FITS[[1]])

par(mfrow = c(1,2))

plot(RD, col.level = NA, col.grid = NA, ext = EXT)  # range distr doesn't change much but will become less detailed (until closer to Gaussian w/ high uncertainty)
title("Range distribution")
plot(OD, col.level = NA, ext = EXT)  # will lose more and more of the distribution until it's a single point
title("Occurrence distribution")

```

```{r sampling_duration3}

# Remove the second half of the data
SUB <- SUB[1:round(nrow(SUB)/2),]  # cutting data in half

RD <- akde(SUB, FITS[[1]])
OD <- occurrence(SUB, FITS[[1]])

par(mfrow = c(1,2))

plot(RD, col.level = NA, col.grid = NA, ext = EXT)  # range distr doesn't change much but will become less detailed (until closer to Gaussian w/ high uncertainty)
title("Range distribution")
plot(OD, col.level = NA, ext = EXT)  # will lose more and more of the distribution until it's a single point
title("Occurrence distribution")

```

```{r sampling_duration4}

# Remove the second half of the data
SUB <- SUB[1:round(nrow(SUB)/2),]  # cutting data in half

RD <- akde(SUB, FITS[[1]])
OD <- occurrence(SUB, FITS[[1]])

par(mfrow = c(1,2))

plot(RD, col.level = NA, col.grid = NA, ext = EXT)  # range distr doesn't change much but will become less detailed (until closer to Gaussian w/ high uncertainty)
title("Range distribution")
plot(OD, col.level = NA, ext = EXT)  # will lose more and more of the distribution until it's a single point
title("Occurrence distribution")

```

## Final Take-Home Concepts

**Range area** = predicted space use, given the same behaviors (biological)

**Occurrence area** = uncertainty (sampling dependent and limited to the sampling period)

**_Neither estimate the amount of space used during the sampling period!_**

---

## Key functions reference

| Function | Package | Purpose | 
|:---|:---|:---|
| `akde()` | `ctmm` | Calculate a range distribution based on autocorrelated kernel density estimation (AKDE) | 
| `occurrence()` | `ctmm` | Calculate an occurrence distribution using Kriging |
| `extent()` | `ctmm` | Calculate the $(x,y)$ plotting extent (bounding box) of ctmm objects |

---

_Next: **Module 2 Part b — Effective Sample Sizes** — We examine effective sample sizes and how the information content in your data can affect your analysis._

