---
title: "M2-b: Effective Sample Sizes"
author: "Animal Movement Analyses for Ecologists | SYSU Shenzhen, July 2026"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    toc_collapse: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Created: 2026-06-30* | *Last updated: `r Sys.Date()`* 

<a href="https://github.com/Animal-Movements/China_2026.git" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#151513; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

---

## Learning Objectives

By the end of this module you will be able to:

1. Understand the effective and absolute sample sizes.
2. Perform parametric bootstrapping to improve parameter estimates and confidence intervals.

**Packages used:** `ctmm`

---

## Background

What is an "effective sample size" (ESS)?
also called "degrees of freedom" (DOF)
Example: consider the standard error of the IID mean estimate from Statistics 101:
S.E. = S.D./sqrt(n)
S.E. = the standard error (uncertainty) of the mean estimate
S.D. = the standard deviation of the process itself
n    = the nominal sample size (the actual number of data points)
With larger 'n', the standard error (uncertainty) is smaller 
But with autocorrelated data, this exact relationship does not hold
Given S.E., then inverting the relationship S.E. = S.D./sqrt(N)
to get N = S.D.^2/S.E.^2 then 'N' is the "effective sample size"
N is a relative measure of the certainty
For IID data, N=n. The effective sample size is the nominal sample size
For autocorrelatd data, N<n. The effective sample size is less than the nominal sample size
If you had N IID data points, then you would get the same quality estimate

## Data Preparation

Here, we will use tracking data from GPS-collared Mongolian gazelles (*Procapra gutturosa*). 

### Import and Visualize Data

```{r data_import}

library(ctmm)

# load gazelle data
data(gazelle)

# select 18th gazelle
DATA <- gazelle[[18]]

COL <- color(DATA, by = 'time')
plot(DATA, col = COL)
## seasonal nomadic range crossing time
## weak site fidelity
## this particular indiv crosses range 2-3 times

```

### Model Selection

```{r ctmm.select}

# guestimate model parameters
GUESS <- ctmm.guess(DATA,interactive=FALSE)

# select best model
FIT <- ctmm.select(DATA,GUESS,trace=3)
# save("FIT",file="gazelle.rda")
load("Day3_HomeRange/Data/gazelle.rda")

# summarize data
summary(DATA)
# note the sampling period

# summarize model
summary(FIT)
# note the effective sample sizes (DOF for mean and area)
## central location and size of area
## here, approx. 2 effective data points for continuous time, but ~372 for RSF/SSFs

```

## Effective Sample Size

```{r ESS}

# SI units converter
help("%#%")
1 %#% 'hr'

# approximate effective sample size
(12.22181 %#% "month") / (5.741925 %#% "month")  # period of data/HR estimation (should be about the # of HR crossings)

# compare to estimate
summary(FIT)$DOF
## high bias (only 2 effective data points)

# ctmm.fit/ctmm.select help file
help("ctmm.fit")
# note the methods argument
# default is not max likelihood ML, need specific min effective sample sizes to work properly


############################
# For a target bias of O(5%)
# ctmm.fit method="ML" requires DOF[area] >= 20 points       (CONVENTIONAL)
# ctmm.fit method="pHREML" requires DOF[area] >= 4-5 points  (DEFAULT)
# ctmm.boot method="pHREML" requires DOF[area] >= 2-3 points (SLOW)
# but in all cases DOF[area] is an estimate

```

## Bootstrapping

```{r bootstrap}

help("ctmm.boot")

# this will take a while
BOOT <- ctmm.boot(DATA,FIT,cores=-1)
## will take a long time (fitting as many times needed to reach threshold)

# discuss daily, weekly, monthly, seasonal home-range estimates
## timescale of estimates depends on specific species/pop and sampling

```

---

## Key functions reference

| Function | Package | Purpose | 
|:---|:---|:---|
| `ctmm.boot()` | `ctmm` | Perform parametric bootstrapping to improve on point estimates and confidence intervals from an initial estimated movement model (NOTE: computationally expensive) | 

---

_Next: **Module 3 — AKDE, Home Range Meta-Analysis, Population Ranges** — We estimate home ranges using autocorrelated kernel density estimation (AKDE), perform a meta-analysis to compare home ranges across individuals, and estimate population-level range distributions._

