---
title: 'M4-a: Periodic Movement Patterns'
subtitle: "Day 4 — Behavior Analysis Part 1"
author: "Animal Movement Analyses for Ecologists | SYSU Shenzhen, July 2026"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    toc_collapse: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Created: 2026-07-13* | *Last updated: `r Sys.Date()`*

<a href="https://github.com/Animal-Movements/China_2026.git" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#151513; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

---

## Learning Objectives

By the end of this module you will be able to:

1. Visualize periodicity in the autocorrelation structure of telemetry data.
2. Use a periodogram to determine potential periodic patterns of space use.
3. Incorporate periodicity into the movement model fitting process.

**Packages used:** `ctmm`

---

## Background

## Data Preparation

### Import and Visualize Data

```{r import_data}

library(ctmm)

# vignette("variogram")

data(wolf)

DATA <- wolf$Gamba

plot(DATA)

SVF <- variogram(DATA)

plot(SVF, frac = 0.01)  # zoom in

SVF <- variogram(DATA,CI="Gauss")
zoom(SVF)

```

```{r autocorrelation, eval = FALSE}

?acf  # or help(acf)
?correlogram  # or help(correlogram)

```

### Model Selection

```{r ctmm.select}

ctmm.guess(DATA,variogram=SVF)
GUESS <- ctmm.guess(DATA,variogram=SVF,interactive=FALSE)

FITS <- ctmm.select(DATA,GUESS,verbose=TRUE,trace=2)

summary(FITS)
summary(FITS[[1]])
zoom(SVF,CTMM=FITS[[1]])

summary(FITS[[2]])

summary(FITS[[3]])

UD <- akde(DATA,FITS[[1]],weights=TRUE,PC="direct",fast=FALSE,trace=2)
plot(DATA,UD)

FITS$IID <- ctmm.fit(DATA)
UD.IID <- akde(DATA,FITS$IID)
plot(DATA,UD.IID)

summary(UD)
summary(UD.IID)

```

## Periodogram

```{r periodogram, eval = FALSE}

vignette("periodogram")

```

```{r periodogram2}

LSP <- periodogram(DATA,fast=FALSE,res.time=2,res.freq=2)
plot(LSP)
plot(LSP,diagnostic=TRUE,max=TRUE)

```

## Fitting a Periodic Movement Model

```{r periodic_model}

PROTO <- ctmm(mean="periodic")
GUESS <- ctmm.guess(DATA,variogram=SVF,CTMM=PROTO,interactive=FALSE)

PFITS <- ctmm.select(DATA,GUESS,verbose=TRUE,trace=2)

summary(PFITS)

zoom(SVF,CTMM=PFITS[[1]])
zoom(SVF,CTMM=PFITS[[4]])

summary(PFITS[[1]])

names(UD)
UD@info
image(UD$PDF)
image(UD$CDF)
UD$r

```

---

## Key functions reference

| Function | Package | Purpose | 
|:---|:---|:---|
| `variogram()` | `ctmm` | Calculate an empirical variogram from movement data | 
| `periodogram()` | `ctmm` | Calculate the isotropic Lomb-Scargle periodogram of tracking data to visualize periodocities that result from an autocorrelated sampling schedule |
| `ctmm()` | `ctmm` | Fit a hypothetical movement model by specifying the model parameters; only used in this case to specify `mean = "periodic"` for periodicity to be accounted for in the `ctmm.guess` and `ctmm.select` |

---

_Next: **Module 4 Part b — Encounter/Interaction Analysis** — We examine methods for investigating potential interactions and encounters between individuals in a population (or across species)._


