---
title: "M4-b: Encounter/Interaction Analysis"
author: "Animal Movement Analyses for Ecologists | SYSU Shenzhen, July 2026"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    toc_collapse: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Created: 2026-07-13* | *Last updated: `r Sys.Date()`*

<a href="https://github.com/Animal-Movements/China_2026.git" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#151513; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

---

## Learning Objectives

By the end of this module you will be able to:

1. Investigate interactions between individuals using various methods available through `ctmm`:
2. Estimate home range overlap.
3. Calculate pairwise distances and proximity ratios.
4. Estimate encounter rates between individuals and their encounter location distributions (conditional distribution of encounters [CDE])

**Packages used:** `ctmm`

---

## Background

## Data Preparation

```{r import_data}

# These analyses are conditional on fitted movement models and HR estimates
# (see: https://github.com/ctmm-initiative/ctmmlearn/blob/main/ctmm_akde.R)
library(ctmm)

data("buffalo")
projection(buffalo) <- median(buffalo)
load("Data/buffalo.rda") # Fitted movement models; object called 'FITS'
load("Data/buffalo_akdes.rda") # Estimated HR areas; object called 'AKDES'

```

```{r HR_overlap, eval = FALSE}

#-----------------------------------------------------
# Home-range overlap
#-----------------------------------------------------

# Do individuals share the same space?
# Relevant paper: https://doi.org/10.1111/2041-210X.13027
help("overlap")  ## overlap of distributions (measure of similarities btwn distributions as % overlap of PMF)

```

```{r HR_overlap2}

#Estimate HR overlap for all pairs
#Note: these all must have compatible resolutions and alignments
OVER <- overlap(AKDES)  ## statistical measure of overlap
## (as compared to geometric overlap that depends on the % range chosen, usually 95%, and doesn't consider freq of space use)
OVER  ## has bias correction (things tend to look more dissimilar than they are, so overlap estimates are lower w/out bias correction)

# This will generate an error because of incompatible grids (due to pixel by pixel calculation)
overlap(list(akde(buffalo$Pepper, FITS$Pepper),
             akde(buffalo$Queen, FITS$Pepper)))

# But this works because HRs are estimated simultaneously (and consistently)
overlap(akde(list(buffalo$Pepper,buffalo$Queen),
             list(FITS$Pepper, FITS$Queen)))


# look at everything
OVER

# pairwise CIs 
OVER$CI["Pepper","Toni",]
OVER$CI["Queen","Toni",]

# point estimates
OVER$CI[,,"est"]

```

```{r CDE, eval = FALSE}

#-----------------------------------------------------
# Encounter location distributions (CDE)
#-----------------------------------------------------

# where encounters are expected to take place (assumption that indivs are moving independently when not encountering)

# Relevant paper: https://doi.org/10.1111/2041-210X.13597
help("cde")  # estimated area of where animals are likely to encounter each other

```

```{r cde2}

#Plot the data and HR estimates
plot(buffalo[c("Pepper", "Queen")],
     UD=AKDES[c("Pepper", "Queen")],
     col = c("#e76f51", "#264653"),
     col.DF=c("#f4a261", "#2a9d8f"),
     col.grid = NA)


#Estimate the home range overlap
overlap(AKDES[c("Pepper", "Queen")])


#Estimate the CDE (conditional distribution of encounters, where we expect majority of encounter to takep place)
CDE <- cde(AKDES[c("Pepper", "Queen")])  ## can weight indivs separately

#Visualise the CDE
plot(buffalo[c("Pepper", "Queen")],
     col=c("#e76f51", "#264653"),
     UD=CDE,
     col.DF="#046C9A",
     col.grid = NA)

```

```{r distance, eval = FALSE}

#-----------------------------------------------------
# Pairwise proximity and distance metrics
#-----------------------------------------------------

# metrics that takes time into account (paper coming)
help("proximity")

```

```{r distance2}

#Pairwise separation distances
DISTS <- distances(buffalo[c("Cilla","Mvubu")],
                   FITS[c("Cilla","Mvubu")])
## predicts distance btwn 2 indivs at given time, might have location


#Visualise the separation distances
plot(DISTS$est ~ DISTS$timestamp,
     type = "l",
     col = "#5e548e")


# Internal plotting function (work in progress)
ctmm:::ts.plot(DISTS)

# what would totally independent motion look like?
cilla_sim <- simulate(FITS$Cilla, t = buffalo$Cilla$t)
mvubu_sim <- simulate(FITS$Mvubu, t = buffalo$Mvubu$t)

sim_dists <- distances(list(cilla_sim, mvubu_sim),
                       FITS[c("Cilla","Mvubu")])

#Plot the data
par(mfrow = c(2,2))
plot(buffalo[c("Cilla", "Mvubu")],
     col = c("#e76f51", "#264653"),
     main = "Empirical data")

plot(list(cilla_sim, mvubu_sim),
     col = c("#e76f51", "#264653"),
     main = "Simulated data")

plot(DISTS$est ~ DISTS$timestamp,
     type = "l",
     col = "#5e548e",
     main = "Empirical distances",
     ylab = "Distance (m)",
     xlab = "Time",
     ylim = c(0,max(sim_dists$est)))

plot(sim_dists$est ~ sim_dists$timestamp,
     type = "l",
     col = "#5e548e",
     main = "Simulated distances",
     ylab = "Distance (m)",
     xlab = "Time",
     ylim = c(0,max(sim_dists$est)))

```

```{r proximity, eval = FALSE}

# Proximity ratio (note: can be slow)
help('proximity')

```

```{r proximity2}

PROXIMITY <- proximity(buffalo[c("Cilla","Mvubu")],
                       FITS[c("Cilla","Mvubu")])
load("Day4_BehaviourAnalysis_PT1/Data/buffalo_proximity.rda")  # ond statistic only works if they are moving tog or actually avoiding each
PROXIMITY  ## >1 = farther apart than each other <1 closer together than expected

# Proximity ratio for simulated animals
SIM_PROXIMITY <- proximity(list(cilla_sim, mvubu_sim),
                           FITS[c("Cilla","Mvubu")])
load("Day4_BehaviourAnalysis_PT1/Data/simulated_proximity.rda")
SIM_PROXIMITY

```

```{r encounter, eval = FALSE}

#-----------------------------------------------------
# Encounters
#-----------------------------------------------------

help("encounter")  ## encounters, encounter rates/freqs
# Relevant paper: https://doi.org/10.1101/2023.06.07.544097
## old measure is prox, which is very sensitive (looks at whether they're w/in a certain threshold distance of each other)
## but the issue is that this misses a lot of potential encounters if data is too coarse.

```

```{r encounter2}

#Empirical encounters
DISTS$encounter <- ifelse(DISTS$est <= 100, 1, 0)

#Visualize the results
par(mfrow = c(1,1))
plot(DISTS$encounter ~ DISTS$timestamp)
cdplot(as.factor(DISTS$encounter) ~ DISTS$timestamp)

#Empirical Encounter rate (n/day)
n <- sum(DISTS$encounter)
t <- "day" %#% (DISTS$t[nrow(DISTS)] - DISTS$t[1])
n/t


#If you do this, run a sensitivity analysis
enc_rad <- 1:1000
N <- vector("numeric", 1000)
for(i in 1:length(enc_rad)){
  N[i] <- sum(ifelse(DISTS$est <= enc_rad[i], 1, 0))
}

#visualise the results
plot(N ~ enc_rad,
     ylab = "Encounters",
     xlab = "Encounter radius",
     type = "l",
     col = "#5e548e")


#Estimate relative encounter rates
RATES <- encounter(AKDES, method = "PDF")
RATES$CI["Cilla","Mvubu",] * 1000^2 # good for small distances (spend about 1.2% of their time tog)
tanh(sqrt(RATES$CI["Cilla","Mvubu",])*1000)^2 # more reliable
## strong assumption that when they're not together, they're avoiding each other

```

---

## Key functions reference

| Function | Package | Purpose | 
|:---|:---|:---|
| `overlap()` | `ctmm` | Calculate the overlap between two stationary distributions using the Bhattacharyya coefficient | 
| `cde()` | `ctmm` | Estimate the conditional location distribution of where encounters take place |
| `proximity()` | `ctmm` | Estimate the proximity of two individuals at shared times from their telemetry data and movement models |
| `encounter()` | `ctmm` | Calculate encounter statistics such as standardized encounter probabilities with confidence intervals and encounter rates |

---

_Next: **Module 5 — Path Reconstruction & Simulations** — We predict the most likely path of an individual from its tracking data and selected movement model. We simulate location points and movement models, as well as utilization distributions, conditioned on the data._


