---
title: 'M5: Path Reconstruction & Simulations'
subtitle: "Day 4 — Behavior Analysis Part 1"
author: "Animal Movement Analyses for Ecologists | SYSU Shenzhen, July 2026"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    toc_collapse: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Created: 2026-07-13* | *Last updated: `r Sys.Date()`*

<a href="https://github.com/Animal-Movements/China_2026.git" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#151513; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

---

## Learning Objectives

By the end of this module you will be able to:

1. Make conditional predictions for animal locations based on the telemetry data and selected movement model, to get the most-likely-path taken by an individual.
2. Simulate paths from a defined movement model and sampling schedule, conditional on data (if provided).
3. Include parameter uncertainty in simulations by simulating a movement model from the sampling distribution of a defined movement model, conditional on data (if provided).

**Packages used:** `ctmm`

---

## Background

[INSERT BACKGROUND ON WHAT CAN BE GAINED FROM PATH RECONSTRUCTION AND SIMULATIONS]

## Data Preparation

### Import and Visualize Data

```{r import_data}

# Load ctmm package
library(ctmm)

# Load buffalo data from package
data(buffalo)
projection(buffalo) <- median(buffalo)  # center projection on median

names(buffalo)

# Let's look at Cilla
DATA <- buffalo$Cilla
plot(DATA, col = color(DATA, by = "time"))

```

### Model Selection

```{r ctmm.select, eval = FALSE}

# Autocorrelation model selection
GUESS <- ctmm.guess(DATA, interactive = FALSE)
FIT <- ctmm.select(DATA, GUESS, trace = 3)
save(FIT, file = "Data/cilla.rda")

```

```{r ctmm.select}

# I've already run this
load("Data/cilla.rda")

```

## Predicting from the Model

```{r predict}

# Take the first 10 locations (first 10 hrs)
SUB <- DATA[1:10,]

# Plot subset
plot(SUB)

```

```{r predict2, eval = FALSE}

# If working with this amount of data, you might consider alternative ICs
## in particular IC="LOOCV"
help("ctmm.select")
## Just sub-setting for purposes of visualization and speed

```

```{r predict3}

# Convenient conversion function (help("%#%"))
1 %#% 'hr'

# Make an array of times over the same period, but 5 min apart
SEQ <- seq(from=SUB$t[1],to=SUB$t[10],by=5 %#% 'min')

```

```{r predict4, eval = FALSE}

help('predict.ctmm')  # makes conditional preds based on movement and location data

```

```{r predict5}

# Predict locations at those times
PRED <- predict(SUB, FIT, t = SEQ)  # predict for every 5 mins
## can allow smoothing for better location estimates (for when data points have greater or lesser uncertainty)

# Plot predictions & data
plot(list(PRED, SUB), col = c('blue','red'))
## uncertainty ellipses (more certain near data points, but also depends on how long to next point)

```

## Simulating Animal Movement

### Generic Simulations

Simulations can be used to guide study design, by defining model parameters for species of interest.

```{r simulate_gen}

# Position autocorrelation timescale (range crossing time)
tau_p <- 1 %#% 'day'

# Velocity autocorrelation timescale (directional persistence)
tau_v <- 1 %#% 'min'

# Spatial variance
## Note the same as the 95% HR area
## 95% HR area = -2*log(0.05)*pi*sig
sig <- 1 %#% 'kilometers^2'


# Specify an OUF model for simulation
MODEL <- ctmm(tau = c(tau_p, tau_v), isotropic = TRUE, sigma = sig)

# Define the sampling duration/schedule
t <- seq(0,1 %#% 'month', 1 %#% 'hr')

# Simulate from the defined model and sampling schedule
SIM <- simulate(MODEL, t = t)

# Fit the movement models
GUESS <- ctmm.guess(SIM, interactive = FALSE)

FITS <- ctmm.select(SIM, GUESS, cores = -1)

# Summary of the fit (NOTE the DOFs)
summary(FITS)

# Define the sampling duration/schedule
t <- seq(0,4 %#% 'days', 5 %#% 'min')

# Simulate from the defined model and sampling schedule
SIM <- simulate(MODEL, t = t)

# Fit the movement models
GUESS <-  ctmm.guess(SIM, interactive = FALSE)
FITS <- ctmm.select(SIM, GUESS)

# Summary of the fit (NOTE the DOFs)
summary(FITS)

# Home range estimate
HR <- akde(SIM, FITS)

# Visualise the results
plot(SIM, UD = HR)

```

### Conditional Simulations

Simulations conditioned on a fitted movement model (here, African buffalo data)

```{r simulate, eval = FALSE}

help('simulate.ctmm')  # simulate things freely or conditional on data

```

```{r simulate2}

# 1 minute sequence
SEQ <- seq(from = SUB$t[1], to = SUB$t[10], by = 1 %#% 'min')

# Simulate locations at those time
SIM <- simulate(SUB, FIT, t = SEQ)  # simulate path for every min

# Plot conditional simulation & data
plot(list(SIM, SUB), col = c('blue','red'), type = c('l','p'))

SIM2 <- simulate(SUB, FIT, t = SEQ)

# Plot conditional simulation & data
plot(list(SIM, SIM2, SUB), col = c('blue','orange','red'), type = c('l','l','p'))
## some variation in simulated trajectories based on the data
## variance would match up with the error ellipses

```

```{r emulate, eval = FALSE}

# That is only trajectory uncertainty
## can also include parameter uncertainty
help('emulate')  # simulating movement model itself from posterior of estimate of movement model (for when model is unknown)

```

```{r emulate2}

# Include parameter uncertainty
SIM3 <- simulate(SUB, emulate(FIT, fast = T), t = SEQ)  
## emulate(FIT) for sample movement model to approximate unknown movement model

plot(list(SIM, SIM2, SIM3, SUB), 
     col = c('blue','orange','black','red'), type = c('l','l','l','p'))

```

## Coati Case Study

### Data preparation

```{r coati}

# Load in the dataset from ctmm
data("coati")
summary(coati)

projection(coati) <- median(coati)

# Subset to single individual
DATA <- coati[[1]]

# Visualise the data
plot(DATA)

```

### Model Selection

```{r coati2}

# Generate the variogram
vg <- variogram(DATA)

GUESS <- ctmm.guess(DATA, variogram = vg, interactive = FALSE)

# Visually inspect the fit
plot(vg, GUESS)

# Fit the models
FIT <- ctmm.select(data = DATA, CTMM = GUESS, cores = -1, trace = 3)

# Summary of the fit
summary(FIT)

```

### Simulating Paths

```{r coati3}

# Simulate from the fitted model, with the original sampling times
SIM <- simulate(FIT, t = DATA$t)

# Visualise the results
plot(list(DATA, SIM), col = c("red", "blue"))

# Draw model parameters from the sampling distribution of the fitted model
FIT2 <- emulate(FIT, fast = TRUE)

```

We can generate many simulations to see potential paths taken under the same movement behaviors and conditions.

```{r coati4}

# Simulate from the model
SIM2 <- simulate(FIT2, t = DATA$t)

# Visualise the results
plot(list(DATA, SIM, SIM2), col = c("red", "blue", "grey30"))

# Fill in the gaps 
SIM3 <- simulate(DATA, CTMM = FIT, dt = 60)

# Visualise the results
plot(list(DATA, SIM3), type = c("p", "l"), col = c("red", "blue"))

```

### Predict the Most-Likely Path

```{r coati5}

# Generate the most likely path (conditioned off of the fitted model)
MLP <- predict(DATA, FIT, dt = 60)

plot(DATA)
plot(MLP, error = FALSE, type = "l", col = "blue", add = TRUE)

# Export the simulations to work with other packages
SIM_EXPORT <- simulate(DATA, CTMM = FIT, complete = TRUE)

EXPORT <- SpatialPoints.telemetry(SIM_EXPORT)

class(EXPORT)
plot(EXPORT)

```

---

## Key functions reference

| Function | Package | Purpose | 
|:---|:---|:---|
| `predict.ctmm()` or `predict()` | `ctmm` | Predict animal locations from a continuous-time movement model | 
| `simulate()` | `ctmm` | Simulate animal locations and tracks from a continuous-time movement model (and conditioned on data, if provided) |
| `emulate()` | `ctmm` | Generate random model-fit statistics from the sampling distribution of a continuous-time movement model and given sampling schedule |

---

_Next: **Day 5 — Behavior Analysis Part 2** — We will continue our exploration of movement behaviors through behavioral classification with Hidden Markov Models (HMM). We will also walk through how remote sensing data can be incorporated into our analyses, including how to account for issues remotely-sensed data, such as GPS location error._

